---
title: "Soft community detection in networks with nmfkc.net"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soft community detection in networks with nmfkc.net}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

`nmfkc.net()` is a **symmetric** NMF for network (adjacency) matrices.  With
`type = "bi"` it factorises a symmetric \eqn{N \times N} matrix as
\eqn{Y \approx X X^\top} with \eqn{X \ge 0}.  Each row of \eqn{X} is a node's
**graded membership** across communities, so this is *soft* community detection:
unlike a hard method (e.g.\ Louvain), boundary nodes are revealed by a mixed
membership rather than being forced into one group.

This vignette uses a **self-contained synthetic network with a known community
structure**, so we can check that `nmfkc.net()` recovers it and that
`nmfkc.net.rank()` finds the right number of communities.

```{r lib}
library(nmfkc)
```

## A synthetic network (stochastic block model, 3 communities)

We draw an undirected graph on `60` nodes from a **stochastic block model**:
three communities of 20 nodes, with a high within-community edge probability and
a low between-community one.  The true community of each node is known.

```{r data, fig.width = 6, fig.height = 5.2}
set.seed(1)
K <- 3; n_each <- 20; N <- K * n_each       # 60 nodes, 3 communities
block <- rep(1:K, each = n_each)            # true community labels

p_in <- 0.6; p_out <- 0.05
Prob <- matrix(p_out, N, N)
for (k in 1:K) Prob[block == k, block == k] <- p_in
Y <- matrix(rbinom(N * N, 1, Prob), N, N)   # 0/1 adjacency
Y[lower.tri(Y)] <- t(Y)[lower.tri(Y)]        # make symmetric
diag(Y) <- 0

isSymmetric(Y); sum(Y) / 2                   # symmetric, number of edges

# adjacency matrix in the original (random) node order -- structure is hidden
image(1:N, 1:N, Y[, N:1], col = c("white", "steelblue"),
      xlab = "node", ylab = "node", main = "Adjacency (original order)")
```

## Soft community detection with `nmfkc.net`

```{r fit}
res <- nmfkc.net(Y, rank = 3, type = "bi", nstart = 10, maxit = 500)
res$r.squared                 # fit
head(round(res$X.prob, 2))    # soft membership: rows = nodes, columns = communities
head(res$X.cluster)           # hard label = argmax membership
```

## Did it recover the known communities?

```{r compare}
table(true = block, estimated = res$X.cluster)
```

The cross-tabulation is (block-)diagonal: every node is assigned to its true
community.

## Visualising the result

Re-ordering the nodes by the **recovered** community turns the adjacency matrix
into a clean block-diagonal, and the membership matrix shows three crisp blocks:

```{r viz, fig.width = 8, fig.height = 4}
ord <- order(res$X.cluster)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
image(1:N, 1:N, Y[ord, rev(ord)], col = c("white", "steelblue"),
      xlab = "node (reordered)", ylab = "", main = "Adjacency by community")
image(1:N, 1:K, res$X.prob[ord, , drop = FALSE],
      col = hcl.colors(20, "Blues", rev = TRUE),
      xlab = "node (reordered)", ylab = "community", axes = FALSE,
      main = "Soft membership")
axis(1); axis(2, at = 1:K)
par(op)
```

## Network diagram (Graphviz DOT)

`nmfkc.net.DOT()` turns the fit into a Graphviz **DOT** graph: each network node
is linked to the community (basis) it belongs to, the edge thickness encodes the
membership strength, and colours mark the communities.  Rendered with the
suggested package **DiagrammeR**, it embeds directly in the HTML output.

```{r dot-check, include = FALSE}
has_dg <- requireNamespace("DiagrammeR", quietly = TRUE)
```

```{r dot, eval = has_dg, fig.width = 7, fig.height = 7}
dot <- nmfkc.net.DOT(
  res,
  Y.label = as.character(1:N),
  X.label = paste("Community", 1:K),
  Y.title = "Network nodes",
  X.title = "Communities",
  layout  = "fdp"            # force-directed; try "circo", "neato", "dot"
)
DiagrammeR::grViz(as.character(dot))
```

Each node is pulled toward the community hub it belongs to; a node with a split
membership would sit between two hubs.  `Y.cluster = "hard"` colours every node by
its single dominant community instead.  (If DiagrammeR is not installed this
diagram is skipped.)

## How many communities? `nmfkc.net.rank`

The same rank-selection diagnostics as `nmfkc.rank` are available for the
symmetric model.  The recommended rank is the cross-validation (ECV) minimum;
the effective rank and R-squared elbow corroborate it.

```{r rank, fig.width = 7, fig.height = 6}
rk <- nmfkc.net.rank(Y, rank = 1:6, type = "bi", nstart = 5)
rk$rank.best
round(rk$criteria, 3)
```

## Remarks

`nmfkc.net()` recovers the three planted communities exactly, and
`nmfkc.net.rank()` selects three.  Because the membership is *soft*
(`res$X.prob`), a node that sits between communities would show a split
membership instead of being forced into one group --- the practical advantage of
NMF over hard community detection.

`type = "bi"` (\eqn{Y \approx X X^\top}) is the usual choice for an undirected,
non-negative network; `type = "tri"` (\eqn{Y \approx X C X^\top}) adds a
community-interaction matrix \eqn{C}, and `type = "signed"` handles networks with
negative weights.  For choosing the number of communities, the same advice as in
the *rank-selection* vignette applies: use the ECV minimum as the primary
criterion and corroborate it with the effective rank.

See `?nmfkc.net`, `?nmfkc.net.ecv`, `?nmfkc.net.rank`, and the companion vignette
*"Choosing the NMF rank on data with a known true rank"*.
