---
title: "NMF-GMM: Covariate-Adjusted Model-Based Clustering with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-GMM: Covariate-Adjusted Model-Based Clustering with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{MASS}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Introduction

NMF with covariates approximates the data by $Y\approx X\Theta A = XB$: a
non-negative basis $X$ and covariate-driven scores $B=\Theta A$. NMF-RE
(`nmfre()`) adds a single Gaussian random effect to the scores.
**NMF-GMM** goes one step further and places a $K$-component *Gaussian
mixture* on the scores,
$$\boldsymbol b_n\mid(z_n{=}k)\sim N_Q(C\boldsymbol a_n+\boldsymbol\mu_k,\Sigma_k),\qquad
  \boldsymbol y_n = X\boldsymbol b_n+\boldsymbol\varepsilon_n,\quad X\ge 0,$$
so that regression (the covariate term $C\boldsymbol a_n$, with $C=\Theta$) and
model-based clustering (the mixture, through the posterior responsibilities) are
fitted **at once**. $K=1$ is NMF-RE.

Why fit them jointly? Because *adjusting first and clustering afterwards* can
displace the class means by an omitted-variable bias --- the omitted variable
being the class. Keeping the covariate **and** the mixture lets one ask whether
latent subpopulations persist *after* adjusting for a covariate. This vignette
shows the effect on the Leptograpsus **crabs** data, where overall size is
confounded with the four species--sex groups.

As elsewhere in the package, `nmf.gmm()` performs optimization only; use
`nmf.gmm.inference()` for standard errors of $C$ and `nmf.gmm.select()` to
choose $K$.

---

## 1. Data: Leptograpsus crabs

The `crabs` data (`MASS`) holds five morphological measurements on 200 crabs:
two species (blue / orange) $\times$ two sexes, 50 each. The rows of $Y$
are the five measurements; the columns are crabs. A natural nuisance covariate
is overall **size** (the average measurement), which is strongly confounded with
the four groups.

```{r data}
library(nmfkc)
library(MASS)
data(crabs)

meas <- c("FL", "RW", "CL", "CW", "BD")
Y    <- t(as.matrix(crabs[, meas]))                 # 5 x 200 (non-negative)
grp  <- interaction(crabs$sp, crabs$sex)            # 4 species-sex groups
crab_df <- data.frame(size = rowMeans(crabs[, meas]))  # one row per crab

## Warm-start the mixture from a plain rank-3 NMF basis. Anchoring every fit
## on the same non-negative parts makes the runs reproducible and comparable,
## and keeps the score space fixed across the with/without-covariate contrast.
X0   <- nmfkc(Y, rank = 3, seed = 1)$X
```

---

## 2. Clustering without adjusting for size

With no covariate (the default `A = NULL` supplies the intercept row), NMF-GMM
is a Gaussian mixture on
the raw scores. Size dominates the leading score directions, so the mixture
tends to split the crabs by **size**, not by species/sex.

```{r fit-nocov}
fit_a <- nmf.gmm(Y, rank = 3, K = 4, cov = "tied", X.init = X0, seed = 1)
ari <- function(cl) round(nmfkc:::.nmfgmm.ARI(cl, grp), 3)
ari(fit_a$cluster)      # adjusted Rand index vs the four species-sex groups
```

The recovery of the four groups is poor: the clustering is largely picking up
the size gradient.

---

## 3. Clustering *while* adjusting for size

Now pass size as a covariate. `A` accepts a one-sided **formula** evaluated in
`data`: the design matrix is built with `model.matrix()`, the intercept row is
added, and the covariate columns are centered and scaled by default (a factor
covariate would expand to centered indicator rows the same way). The covariate
term
$C\boldsymbol a_n$ absorbs the size effect on the scores, and the mixture clusters
the **size-adjusted** structure --- recovering the species--sex groups.

```{r fit-cov}
fit_b <- nmf.gmm(Y, ~ size, data = crab_df, rank = 3, K = 4, cov = "tied",
                 X.init = X0, seed = 1)
ari(fit_b$cluster)
```

Adjusting for size *inside* the mixture lifts the adjusted Rand index sharply.
The confusion table against the true groups shows nearly clean recovery:

```{r confusion}
table(class = fit_b$cluster, group = grp)
```

`print()` / `summary()` give the model-based clustering report (mixing
proportions, cluster sizes, BIC/ICL):

```{r print}
fit_b
```

---

## 4. The matched two-stage baseline: `nmf.gmm.twostage()`

The obvious alternative to adjusting *while* clustering is adjusting *first*:
regress the scores on size, then cluster the residuals.
`nmf.gmm.twostage()` packages that recipe with everything held equal to the
joint fit --- same basis initialization, same mixture, same settings --- so
that **only the order** of adjustment and clustering differs:

```{r twostage}
fit_c <- nmf.gmm.twostage(Y, ~ size, data = crab_df, rank = 3, K = 4,
                          cov = "tied", X.init = X0, seed = 1)
c(two.stage = ari(fit_c$cluster), joint = ari(fit_b$cluster))
```

The two-stage route falls short of the joint fit, and in the predicted
direction: fitted blind to the class, the stage-one regression charges to size
part of what separates the species--sex groups, so the residuals it hands to
the mixture have had some of the class contrast subtracted along with the
nuisance (Satoh 2026, Proposition 4). When the covariate is
(near-)mean-independent of the class --- for instance under a balanced crossed
design --- the displacement vanishes and the two routes agree; the gap seen
here is bought by the size--class association in the crabs.

---

## 5. Choosing the number of components: `nmf.gmm.select()`

`nmf.gmm.select()` sweeps $K$ and reports BIC / ICL (and, given known
labels, the ARI). On crabs the information criteria mildly **over-select**
(they prefer $K=5$), while clustering quality (ARI) peaks at the true
$K=4$ --- a documented feature of BIC/ICL on real, slightly non-Gaussian
class structure.

```{r select}
sel <- nmf.gmm.select(Y, A = fit_b$A, rank = 3, K = 1:5, cov = "tied",
                      X.init = X0, truth = grp, verbose = FALSE)
sel
```

---

## 6. Inference on the covariate effect: `nmf.gmm.inference()`

`nmf.gmm.inference()` reports Wald inference for the covariate-coefficient
matrix $C$ ($=\Theta$), conditional on the estimated basis and mixture,
with an outer-product mixture-information standard error and a wild-bootstrap
standard error / interval. Each `size` row is the effect of size on the score
of one non-negative part.

```{r inference}
fit_b <- nmf.gmm.inference(fit_b, Y, wild.B = 500, seed = 1)   # A defaults to fit_b$A
subset(fit_b$coefficients, Covariate == "size")
```

The size coefficients are strongly positive on every part (larger crabs load
more on all parts), confirming that size is exactly the nuisance the covariate
term is meant to remove before clustering.

---

## 7. Visualizing the effect of adjustment

Projecting the **covariate-adjusted** scores $\boldsymbol b_n - C\boldsymbol a_n$
onto their first two principal components, and colouring by the true group,
makes the mechanism visible: without adjustment (left) the groups smear along a
size axis; after adjustment (right) they separate.

Read the right-hand panel carefully, because the axes are not the scores
themselves. The model puts the covariate effect in the component means, so the
space the mixture acts in is the score with $C\boldsymbol a_n$ removed --- and
that is what `plot()` draws (its axis labels say `adjusted`). The distinction is
not cosmetic here:

```{r scoreplot-sep}
sep2 <- function(M, g) {                     # between/within on the first 2 PCs
  pc <- prcomp(t(M))$x[, 1:2, drop = FALSE]
  gm <- do.call(rbind, lapply(split(as.data.frame(pc), g), colMeans))
  B  <- sum(table(g) * rowSums((gm - rep(colMeans(pc), each = nrow(gm)))^2))
  W  <- sum(vapply(split(as.data.frame(pc), g), function(d)
         sum((as.matrix(d) - rep(colMeans(d), each = nrow(d)))^2), numeric(1)))
  B / W
}
rbind(raw      = c(a = sep2(fit_a$scores, grp), b = sep2(fit_b$scores, grp)),
      adjusted = c(a = sep2(fit_a$scores - fit_a$C %*% fit_a$A, grp),
                   b = sep2(fit_b$scores - fit_b$C %*% fit_b$A, grp)))
```

The **raw** scores of the size-adjusted fit separate the four groups no better
than those of the plain fit (0.64 either way). The vivid separation in the
right-hand panel --- 4.63 --- is produced by the adjustment. In (a) the two rows
agree exactly, because with no covariate $C\boldsymbol a_n$ is a constant and
subtracting it only shifts the cloud.

So the figure shows *where the mixture is working*, not that the scores
themselves pull apart. The clustering gain is real and is measured by the ARI in
Sections 3--4; the figure explains the mechanism behind it.

The `plot()` method with `type = "adjusted.scores"` draws this directly
(colouring by the true `group` here rather than the fitted cluster):

```{r scoreplot, fig.width=8, fig.height=4, fig.alt="Two adjusted-score scatterplots coloured by the four crab groups: without covariate the groups smear along a size gradient; after adjusting for size they separate into four clusters."}
op <- par(mfrow = c(1, 2))
plot(fit_a, type = "adjusted.scores", group = grp, main = "(a) no covariate")
plot(fit_b, type = "adjusted.scores", group = grp, main = "(b) size-adjusted")
par(op)
```

---

## Summary

| Step | Function | Purpose |
|:---|:---|:---|
| 1 | `nmf.gmm()` | Fit the covariate-adjusted score mixture (optimization only; `A` may be a formula with `data`) |
| 2 | `nmf.gmm.twostage()` | The matched adjust-then-cluster baseline (same basis, same mixture, order reversed) |
| 3 | `nmf.gmm.inference()` | SEs / tests for the covariate coefficients $C$ |
| 4 | `nmf.gmm.select()` | Choose $K$ by BIC / ICL (and ARI if labels are known) |
| — | `predict()`, `summary()`, `plot()` | Hard classes / responsibilities, report, EM convergence |

**The message:** clustering *while* adjusting for a confounder (here, size)
recovers the structure of interest (species and sex) that clustering the raw
scores misses --- the adjusted Rand index rises from about 0.25 to about 0.86,
and beats the matched two-stage pipeline, whose stage-one regression removes
part of the class contrast along with the nuisance.
The covariance structure is controlled by `cov` (`"tied"`, `"free"`, or the
isotropic `"scalar"`, which at $K=1$ is the `nmfre()` model).

For methodological details see Satoh, K. (2026), *NMF-GMM: A Gaussian-Mixture
Latent-Class Extension of Non-negative Matrix Factorization with Covariates*.
