---
title: "NMF-RE: Mixed-Effects Modeling with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-RE: Mixed-Effects Modeling with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{nlme}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates **NMF-RE** (Non-negative Matrix Factorization with Random Effects), a mixed-effects extension of NMF implemented in the `nmfkc` package.

### Why Mixed Effects?

Standard NMF with covariates models the data as:
$$Y \approx X \Theta A$$

where $\Theta A$ captures systematic (fixed) effects of covariates $A$ on latent scores. However, in many applications --- such as longitudinal studies, panel data, or clustered observations --- individuals exhibit **unit-specific deviations** that cannot be explained by covariates alone.

NMF-RE addresses this by adding **random effects** $U$:
$$Y = X(\Theta A + U) + \mathcal{E}$$

- $Y$ ($P \times N$): Non-negative observation matrix (e.g., repeated measurements over $P$ time points for $N$ subjects).
- $X$ ($P \times Q$): Non-negative basis matrix learned from data.
- $\Theta$ ($Q \times K$): Coefficient matrix for covariate effects.
- $A$ ($K \times N$): Covariate matrix (e.g., intercept, sex, treatment).
- $U$ ($Q \times N$): Random effects matrix capturing individual deviations.
- $\mathcal{E}$: Residual noise with variance $\sigma^2$.

The random effects $U$ follow a ridge-type prior controlled by variance $\tau^2$, and the penalty $\lambda = \sigma^2 / \tau^2$ determines the degree of shrinkage.

### Key Features

- **Automatic regularization**: `dfU.cap.rate` controls the effective degrees of freedom consumed by $U$, preventing overfitting.
- **Wild bootstrap inference**: Standard errors, z-values, p-values, and confidence intervals for $\Theta$ without repeated constrained re-optimization.
- **Trace-based ICC**: Quantifies the proportion of variance attributable to random effects.

---

## 1. Data: Orthodont (Longitudinal Growth)

We use the `Orthodont` dataset from the `nlme` package: orthodontic distance measurements for 27 children at ages 8, 10, 12, and 14. The covariate of interest is sex (Male/Female).

```{r load-data}
library(nmfkc)
library(nlme)

data(Orthodont)
head(Orthodont)
```

### 1.1 Prepare the Observation Matrix Y

Each column of $Y$ is a subject, each row is a time point (age). Since there are 4 ages and 27 subjects, $Y$ is $4 \times 27$.

```{r prepare-Y}
Y <- matrix(Orthodont$distance, nrow = 4, ncol = 27)
colnames(Y) <- paste0("S", 1:27)
rownames(Y) <- paste("Age", c(8, 10, 12, 14))
Y[, 1:6]
```

### 1.2 Prepare the Covariate Matrix A

We create a $2 \times 27$ covariate matrix with an intercept row and a binary `male` indicator.

```{r prepare-A}
male <- ifelse(Orthodont$Sex[seq(1, 108, 4)] == "Male", 1, 0)
A <- rbind(intercept = 1, male = male)
A[, 1:6]
```

---

## 2. Selecting dfU Cap Rate

Before fitting the model, we use `nmfre.dfU.scan()` to scan different regularization levels. The **dfU cap rate** limits how many effective degrees of freedom the random effects $U$ can consume, preventing overfitting.

The scan evaluates rates from 0.1 to 1.0 (where 1.0 means no cap). Key columns in the output:

- **dfU**: Effective degrees of freedom actually used.
- **safeguard**: Whether `dfU` stayed below the cap (`TRUE` = safe).
- **hit**: Whether $\tau^2$ was updated freely (`TRUE` = unconstrained).
- **ICC**: Trace-based Intraclass Correlation Coefficient.

```{r dfU-scan}
scan_result <- nmfre.dfU.scan(1:10/10, Y, A, Q = 1)
scan_result
```

### Interpreting the Scan

A good `dfU.cap.rate` satisfies two criteria:

1. **safeguard = TRUE**: The cap is not overly binding.
2. **hit = TRUE**: The model converges to a natural solution without the cap forcing an artificial penalty.

Choose the smallest rate where both conditions hold. This balances model flexibility against overfitting.

---

## 3. Fitting the NMF-RE Model

Based on the scan, we select `dfU.cap.rate = 0.2` and fit the model with rank $Q = 1$ (a single latent trend). The `prefix` argument labels the basis.

```{r fit-nmfre}
res <- nmfre(Y, A, Q = 1, dfU.cap.rate = 0.2, prefix = "Trend")
```

### 3.1 Model Summary

`summary()` displays variance components, fit statistics, and the coefficient table with significance stars.

```{r summary}
summary(res)
```

### Understanding the Output

**Variance components:**

- $\sigma^2$: Residual variance (measurement noise).
- $\tau^2$: Random effect variance (individual heterogeneity).
- **ICC**: Proportion of variance from random effects. Higher ICC means stronger individual differences relative to noise.

**Coefficient table ($\Theta$):**

- **Estimate**: Effect of each covariate on the latent trend.
- **Std. Error**: From wild bootstrap.
- **Pr(>|z|)**: One-sided p-value (H0: $\Theta_{qk} = 0$ vs H1: $\Theta_{qk} > 0$), appropriate because $\Theta$ is non-negative.

**R-squared:**

- $R^2$ (XB+blup): Fit including both fixed and random effects.
- $R^2$ (XB): Fit from fixed effects only --- the gap shows how much random effects improve the fit.

---

## 4. Visualization

We plot the growth curves to see how the model captures both population-level trends (fixed effects) and individual deviations (random effects).

```{r plot-growth, fig.height=6}
age <- c(8, 10, 12, 14)

plot(age, res$XB[, 1], type = "n", ylim = range(Y),
     xlab = "Age (years)", ylab = "Distance (mm)",
     main = "Orthodont: NMF-RE Growth Curves")

# Plot observed data points
for (j in 1:27) {
  pch_j <- ifelse(male[j] == 1, 4, 1)
  points(age, Y[, j], pch = pch_j, col = "gray60")
}

# Plot individual predictions (fixed + random effects)
for (j in 1:27) {
  lines(age, res$XB.blup[, j], col = "steelblue", lty = 3, lwd = 0.8)
}

# Plot population-level fixed effects (two lines: male and female)
for (j in 1:27) {
  lines(age, res$XB[, j], col = "red", lwd = 2)
}

legend("topleft",
  legend = c("Fixed effect (male/female)", "Fixed + Random (BLUP)",
             "Male (observed)", "Female (observed)"),
  lwd = c(2, 1, NA, NA), lty = c(1, 3, NA, NA),
  pch = c(NA, NA, 4, 1),
  col = c("red", "steelblue", "gray60", "gray60"),
  cex = 0.85)
```

### Interpreting the Plot

- **Red solid lines**: Population-level fitted values from fixed effects ($X \Theta A$). There are two distinct curves --- one for males (higher) and one for females (lower) --- reflecting the sex effect in $\Theta$.
- **Blue dashed lines**: Individual-level predictions ($X(\Theta A + U)$). Each line is shifted from the population curve by the subject's random effect $U$, capturing personal growth trajectories.
- **Gray points**: Observed measurements. Crosses ($\times$) for males, circles ($\circ$) for females.

---

## 5. Examining Learned Components

### 5.1 Basis Matrix X

The basis $X$ represents the latent temporal pattern shared across all subjects.

```{r basis-X}
cat("Basis X (temporal pattern):\n")
print(round(res$X, 4))
```

Since $Q = 1$, there is one basis vector. Its shape shows how the latent factor manifests across the four ages. The column is normalized to sum to 1.

### 5.2 Coefficient Matrix $\Theta$

$\Theta$ maps covariates to latent scores:

```{r coef-C}
cat("Coefficient matrix (Theta):\n")
print(round(res$C, 4))
```

- `intercept`: Baseline level of the latent trend for females.
- `male`: Additional contribution for males. A significant positive value indicates that males have higher orthodontic distances on average.

### 5.3 Random Effects U

$U$ captures individual deviations. Subjects with positive $U$ values grow faster than the population average; negative values indicate slower growth.

```{r random-effects, fig.height=4}
barplot(res$U[1, ], names.arg = colnames(Y),
        las = 2, cex.names = 0.7,
        col = ifelse(male == 1, "steelblue", "salmon"),
        main = "Random Effects (U) by Subject",
        ylab = "Random effect value")
legend("topright", fill = c("steelblue", "salmon"),
       legend = c("Male", "Female"), cex = 0.85)
```

---

## 6. Model Diagnostics

### 6.1 Convergence

Check that the algorithm converged properly:

```{r convergence}
cat("Converged:", res$converged, "\n")
cat("Iterations:", res$iter, "\n")
cat("Stop reason:", res$stop.reason, "\n")
```

```{r plot-convergence, fig.height=4}
plot(res$objfunc.iter, type = "l",
     xlab = "Iteration", ylab = "Objective function",
     main = "Convergence Trace")
```

### 6.2 Residual Analysis

Compare the fitted values against the original data:

```{r residual-analysis, fig.height=4}
residuals <- Y - res$XB.blup
cat("Mean absolute residual (BLUP):", round(mean(abs(residuals)), 4), "\n")
cat("Mean absolute residual (fixed):", round(mean(abs(Y - res$XB)), 4), "\n")

# Fitted vs Observed
plot(as.vector(Y), as.vector(res$XB.blup),
     xlab = "Observed", ylab = "Fitted (BLUP)",
     main = "Observed vs Fitted", pch = 16, col = "steelblue")
abline(0, 1, col = "red", lwd = 2)
```

---

## 7. Comparison: With and Without Random Effects

To appreciate the value of random effects, compare NMF-RE with a standard NMF covariate model (no random effects).

```{r compare-models}
# Standard NMF with covariates (no random effects)
res_fixed <- nmfkc(Y, A = A, Q = 1)

cat("=== Standard NMF (fixed effects only) ===\n")
cat("R-squared:", round(1 - sum((Y - res_fixed$XB)^2) / sum((Y - mean(Y))^2), 4), "\n\n")

cat("=== NMF-RE (fixed + random effects) ===\n")
cat("R-squared (XB):      ", round(res$r.squared.fixed, 4), "\n")
cat("R-squared (XB+blup): ", round(res$r.squared, 4), "\n")
cat("ICC:                 ", round(res$ICC, 4), "\n")
```

The improvement from fixed-only $R^2$ to BLUP $R^2$ quantifies the contribution of individual random effects.

---

## Summary

NMF-RE provides a principled way to model **individual heterogeneity** in NMF:

| Step | Function | Purpose |
|:---|:---|:---|
| 1 | `nmfre.dfU.scan()` | Select regularization level |
| 2 | `nmfre()` | Fit the mixed-effects model |
| 3 | `summary()` | Examine variance components and inference |

**When to use NMF-RE:**

- Longitudinal or repeated-measures data where subjects differ systematically.
- Panel data with unit-level heterogeneity.
- Any NMF application where fixed covariates alone leave structured residual variation.

For more details on the underlying methodology, see:

- Satoh, K. (2026). Wild Bootstrap Inference for Non-Negative Matrix Factorization with Random Effects. arXiv:2603.01468. <https://arxiv.org/abs/2603.01468>
