---
title: "NMF-SEM with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-SEM with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{lavaan}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette demonstrates how to fit the **NMF-SEM**
(Nonnegative Matrix Factorization–Structural Equation Model)
implemented in the **nmfkc** package.

NMF-SEM extends standard nonnegative matrix factorization by introducing
a **structural equation model in latent space**.

Observed endogenous variables \(Y_1\) are approximated as

\[
Y_1 \approx X (\Theta_1 Y_1 + \Theta_2 Y_2).
\]

If the feedback operator \(XC_1\) is stable, the model implies the equilibrium mapping

\[
\hat Y_1 \approx (I - X\Theta_1)^{-1} X\Theta_2 Y_2
\equiv M_{\mathrm{model}} Y_2.
\]

---

## 1. Load and preprocess data

```{r}
library(lavaan)
data(HolzingerSwineford1939)

d <- HolzingerSwineford1939
index <- complete.cases(d)
d <- d[index, ]
```

### 1.1 Construct exogenous variables

```{r}
d$age.rev <- -(d$ageyr + d$agemo / 12)
d$sex.2 <- ifelse(d$sex == 2, 1, 0)
d$school.GW <- ifelse(d$school == "Grant-White", 1, 0)

d[, c("id", "sex", "ageyr", "agemo", "school", "grade")] <- NULL
```

### 1.2 Nonnegative normalization

```{r}
library(nmfkc)
d <- nmfkc::nmfkc.normalize(d)
```

## 2. Split into endogenous and exogenous blocks

```{r}
exogenous_vars <- c("age.rev", "sex.2", "school.GW")
endogenous_vars <- setdiff(colnames(d), exogenous_vars)

Y1 <- t(d[, endogenous_vars])
Y2 <- t(d[, exogenous_vars])
```

## 3. Baseline NMF model

```{r}
myepsilon <- 1e-6
Q0 <- 3

res0 <- nmfkc(
  Y = Y1,
  A = Y2,
  Q = Q0,
  epsilon = myepsilon,
  X.L2.ortho = 100
)

M.simple <- res0$X %*% res0$C
```

## 4. NMF-SEM estimation

```{r}
p <- list(C1.L1 = 10, C2.L1 = 0.6)

res <- nmf.sem(
  Y1, Y2,
  rank = Q0,
  X.init = res0$X,
  X.L2.ortho = 100,
  C1.L1 = p$C1.L1,
  C2.L1 = p$C2.L1,
  epsilon = myepsilon
)
```

```{r}
plot(res$objfunc.full, type = "l",
     main = "Objective Function",
     ylab = "Loss", xlab = "Iteration")
```

## 5. Diagnostics

```{r}
SC.map <- cor(as.vector(res$M.model),
              as.vector(M.simple))

cat("Q=     ", Q0, "\n")
cat("RHO=   ", round(res$XC1.radius, 3), "\n")
cat("AR=    ", round(res$amplification, 3), "\n")
cat("SCmap= ", round(SC.map, 3), "\n")
cat("SCcov= ", round(res$SC.cov, 3), "\n")
cat("MAE=   ", round(res$MAE, 3), "\n")
```

## 6. Visualization

```{r}
res.dot <- nmf.sem.DOT(
  res,
  weight_scale = 5,
  rankdir = "TB",
  threshold = 0.01,
  fill = FALSE,
  cluster.box = "none"
)

library(DiagrammeR)
grViz(res.dot)
```

## Session information

```{r}
sessionInfo()
```
