---
title: "NMF-FFB with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-FFB with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

NMF-FFB describes non-negative outcomes $Y_1$ ($P_1 \times N$) by a small number
of non-negative factors driven by covariates $Y_2$ ($P_2 \times N$), and asks
whether the outcomes also feed **back** into those factors:

$$
Y_1 = X B + \mathcal{E}, \qquad B = \Theta_1 Y_1 + \Theta_2 Y_2 + U ,
$$

with $X \ge 0$ the basis, $\Theta_2 \ge 0$ the exogenous coefficients,
$\Theta_1 \ge 0$ the feedback, $U \sim N(0, \Phi)$ a unit-level random effect and
$\mathcal{E} \sim N(0, \Psi)$ item noise.  With $A = X\Theta_1$ the equilibrium
is $Y_1 = (I - A)^{-1}(X\Theta_2 Y_2 + XU + \mathcal{E})$, which exists when the
spectral radius $\rho(A) < 1$.

Two facts shape everything below.

**Feedback cannot be found by prediction.**  The equilibrium mapping
$M = (I - X\Theta_1)^{-1}X\Theta_2$ is reproduced exactly by a feed-forward model
with exogenous matrix $L_Q\Theta_2$.  The two models predict $Y_1$ from $Y_2$
identically, so no cross-validation and no test-sample error can prefer one.
What separates them is the *conditional covariance*: with feedback, the
disturbances propagate around the loop.

**An outcome must not feed back into its own factor.**  Without a restriction
the factor-level directions $\Theta_1 = a X^\top \Psi^{-1}$ are observationally
equivalent to a change in $\Theta_2$ and $\Phi$, so they are not identified.  The
default `C1.restriction = "union"` blocks, for each outcome, the factor it loads
on most and every factor on which it loads at least
`C1.restriction.threshold`.

## The workflow

One function per step:

```r
ecv <- nmf.ffb.ecv(Y1, Y2, rank = 1:5)   # 1. choose Q
fit <- nmf.ffb(Y1, Y2, rank = Q)         # 2. estimate; BIC selects the support
tst <- nmf.ffb.test(fit, Y1, Y2)         # 3. test the feed-forward null
dgn <- nmf.ffb.diagnostics(fit)          # 4. cycles, spectral radius, best supports
inf <- nmf.ffb.inference(fit, Y1, Y2)    # 5. intervals for the retained entries
```

The order matters, and step 3 is the only inferential step.  What it tests is

$$
H_0: \text{the distribution of } Y_1 \mid Y_2 \text{ belongs to the feed-forward family}
\qquad\text{against}\qquad
H_1: \text{it does not.}
$$

Note what $H_0$ is *not*.  It is not the single point $\Theta_1 = 0$.  The map
from parameters to distributions is not injective, so the null family is larger
than that point: a feed-forward model with a different $\Theta_2$ and $\Phi$
reproduces the mean *and* the covariance of some models with $\Theta_1 \ne 0$.
Rejection therefore says "no feed-forward model with correlated factors
reproduces these first two moments", not "$\Theta_1 \ne 0$".  Section 1 builds
one data set of each kind so that both outcomes can be seen.  Section 6 explains
why reading step 4 before step 3 is a mistake.

## 1. Data

We use two simulated data sets, so that the truth is known and the two outcomes
of the procedure can both be shown.  Both have $P_1 = 6$ outcomes loading on
$Q = 2$ factors and $P_2 = 3$ covariates.

```{r data}
library(nmfkc)
set.seed(20260909)

N <- 600; P1 <- 6; P2 <- 3; Q <- 2
X.true <- matrix(0.02, P1, Q)                    # parts-based basis: 3 outcomes per factor
X.true[1:3, 1] <- runif(3, 0.7, 1.3)
X.true[4:6, 2] <- runif(3, 0.7, 1.3)
X.true <- sweep(X.true, 2, colSums(X.true), "/")
rownames(X.true) <- paste0("y", 1:P1); colnames(X.true) <- paste0("F", 1:Q)

T2.true <- matrix(0, Q, P2); T2.true[1, 1] <- 1.2; T2.true[2, 2] <- 0.9; T2.true[1, 3] <- 0.5
T1.true <- matrix(0, Q, P1); T1.true[2, 1] <- 0.8; T1.true[1, 4] <- 0.6   # y1 -> F2, y4 -> F1
Y2 <- matrix(runif(P2 * N), P2, N); rownames(Y2) <- paste0("x", 1:P2)

simulate_ffb <- function(T1, feedback.in.loop = TRUE) {
  U <- matrix(rnorm(Q * N, sd = 0.3), Q, N)
  E <- matrix(rnorm(P1 * N, sd = 0.1), P1, N)
  L <- solve(diag(P1) - X.true %*% T1)
  if (feedback.in.loop) {
    Y1 <- L %*% (X.true %*% (T2.true %*% Y2 + U) + E)     # disturbances inside the loop
  } else {
    Y1 <- (L %*% X.true %*% T2.true) %*% Y2 + X.true %*% U + E   # added after it
  }
  Y1 <- pmax(Y1, 0); rownames(Y1) <- rownames(X.true); Y1
}

Y1.H1 <- simulate_ffb(T1.true, feedback.in.loop = TRUE)    # H1 is true
Y1.H0 <- simulate_ffb(T1.true, feedback.in.loop = FALSE)   # H0 is true
```

Both are generated from the *same* $\Theta_1$.  The difference is where the
disturbances enter, and that is what decides which hypothesis holds.

* `Y1.H1`: $U$ and $\mathcal{E}$ enter inside the loop, so they are propagated by
  $(I - X\Theta_1)^{-1}$.  The conditional covariance carries the loop, no
  feed-forward model reproduces it, and $H_1$ is true.
* `Y1.H0`: the same disturbances are added after the loop.  The conditional mean
  is unchanged, but the covariance is $X\Phi X^\top + \Psi$, exactly what the
  feed-forward model with exogenous matrix $L_Q\Theta_2$ gives.  The
  distribution lies in the null family, so $H_0$ is true --- *even though*
  $\Theta_1 \ne 0$ in the code that generated it.

`Y1.H0` is the interesting one: in the parameter space it is a point with
feedback, but as a distribution it is indistinguishable from no feedback.  There
is nothing to detect, and the procedure should say so.

## 2. Choosing the number of factors

```{r ecv}
ecv <- nmf.ffb.ecv(Y1.H1, Y2, rank = 1:4, nfolds = 5, seed = 1)
data.frame(Q = ecv$rank, ECV = round(ecv$objfunc, 4))
Q0 <- ecv$rank[which.min(ecv$objfunc)]
Q0
```

Held-out *entries* of $Y_1$ are predicted from the remaining ones, so this
measures how well a $Q$-factor non-negative basis describes the outcomes -- which
is what $Q$ is for.  It says nothing about feedback, by the first fact above.

## 3. Fitting

`nmf.ffb()` estimates the basis with `nmfkc()`, derives the exclusion
restriction from it, fits the feed-forward null and the full feedback model by
FIML, follows an $L_1$ path, and refits each distinct support without the
penalty; BIC over those supports selects one.

```{r fit}
fit <- nmf.ffb(Y1.H1, Y2, rank = Q0)
round(fit$X, 3)                       # estimated basis
round(fit$C1, 3)                      # selected feedback (Theta1), Q x P1
c(rho = round(fit$XC1.radius, 3), LR = round(fit$LR[["full"]], 2), df = fit$LR.df)
```

The two retained entries should be `y1 -> F2` and `y4 -> F1`, the truth.

The restriction it worked under is on the object, as the rule that produced it
and as the matrix of entries left free:

```{r restriction}
fit$C1.restriction                    # the rule
fit$C1.free                           # 1 = free, 0 = blocked because the outcome loads there
```

Every entry blocked here is one the likelihood is not allowed to use, so the
degrees of freedom of the test are `sum(fit$C1.free)`, not $Q P_1$.

`plot()` draws the BIC path: the line follows the smallest BIC at each penalty,
every proposal is a grey point, the number above each is how many entries that
support keeps, and the circled point is the selected penalty.  There is no
objective trace to plot for a likelihood fit -- its optimizer is L-BFGS-B.

```{r bicpath, fig.alt = "BIC along the L1 path"}
plot(fit)
```

`print(fit)` reports both stages separately, because they are two optimizers:
`iter / maxit` and `epsilon` are the stage-1 NMF that produced the basis, and
`fiml n / m` is the L-BFGS-B of the selected fit.  `converged` is `TRUE` only if
both converged.

```{r print-fit}
fit
```

## 4. Is the feed-forward model enough?

This is the only inferential step.  The likelihood ratio has no $\chi^2$ null
distribution -- $\Theta_1 \ge 0$ puts the null on a boundary and the support is
chosen from the same data -- so it is calibrated by a parametric bootstrap from
the fitted feed-forward null.  The default calibration re-runs **the whole
procedure** on every replicate: the basis is re-estimated, the restriction is
re-derived from it, and the selection follows.

**When $H_1$ is true** the test should reject:

```{r test}
## B = 1000 in practice; kept small here so the vignette builds quickly
tst.H1 <- nmf.ffb.test(fit, Y1.H1, Y2, B = 50, seed = 1)
c(p = round(tst.H1$LR.p.boot[["full"]], 3),
  null.q95 = round(tst.H1$LR.null.quantile[["full"]], 2),
  selects.under.null = tst.H1$prob.select.null,
  restr.moves = round(tst.H1$C1.restriction.change.rate, 2))
```

**When $H_0$ is true** it should not, and BIC should retain nothing --- although
the data were generated with the same non-zero $\Theta_1$:

```{r test-ff}
fit.H0 <- nmf.ffb(Y1.H0, Y2, rank = Q0)
tst.H0 <- nmf.ffb.test(fit.H0, Y1.H0, Y2, B = 50, seed = 1)
c(nnz = sum(fit.H0$C1 > 1e-3),
  p = round(tst.H0$LR.p.boot[["full"]], 3),
  selects.under.null = tst.H0$prob.select.null)
```

Read `prob.select.null` and `C1.restriction.change.rate` alongside the *p*-value.  They say
whether the exclusion restriction is determined well enough for the test to mean
anything: if the dominant factor of an outcome moves from replicate to replicate,
a blocked entry becomes free and a feedback coefficient can absorb loading
structure.  A large `prob.select.null` with a wide null distribution is the
signature of a data set in which the procedure finds feedback whether or not
there is any -- report it and stop, rather than reporting the *p*-value alone.

If the null is not rejected, stop here and report NMF-FF.

## 5. What the feedback looks like

```{r diagnostics}
dgn <- nmf.ffb.diagnostics(fit)
dgn$top                      # the 3 best distinct supports: order, nnz, rho, BIC, dBIC
round(dgn$log.N, 2)          # what one extra free entry costs in BIC
dgn$nested                   # do they form a chain under inclusion?
dgn$core                     # entries common to all three
dgn$cycles$rho               # spectral radius; 0 with a non-empty support = a cascade
dgn$cycles$cycle             # which factors lie on a cycle
```

Adding one free entry costs $\log N$ in BIC, and a difference in BIC is twice the
log of a Bayes factor, so a difference below about 2 is not evidence for one
support over another: the *presence* of feedback can be settled while its
*composition* is not.  When the best supports form a chain under inclusion their
intersection is the part the criterion does not put in doubt; when they do not,
no single support should be read off.

`omega` reports how much of the fitted $\Theta_1$ lies in the factor-level family
that is not identified, against its reference value:

```{r omega}
round(unlist(dgn$omega[c("omega", "omega0")]), 3)
```

## 6. Why the test comes before the spectral radius

The unpenalized $\hat\rho$ is biased upwards under the null, because the
likelihood is nearly flat along the unidentified directions.  The $\hat\rho$ of
the BIC-selected model is not:

```{r rho-order}
c(unpenalized = round(fit.H0$full$XC1.radius, 3),   # on the data where H0 is true
  selected    = round(fit.H0$XC1.radius, 3))
```

Reading $\hat\rho$ first would suggest a loop in data that contain none.  Read it
only after the test has been passed, and treat a positive value as a lower bound:
BIC keeps too few entries to close the loop when the feedback is weak.

## 7. Intervals for the retained entries

After the null has been rejected, `nmf.ffb.inference()` gives bootstrap intervals
for the entries of the *selected* model, with the basis and the support held
fixed.  These describe the sampling variability of the retained coefficients, not
the evidence for retaining them -- an interval that excludes zero is not by itself
a finding.

```{r inference}
inf <- nmf.ffb.inference(fit, Y1.H1, Y2, B = 50, seed = 1)
subset(inf$coefficients, Type == "C1" & Estimate > 0,
       select = c(Basis, Covariate, Estimate, CI_low, CI_high, support_rate))
```

## 8. Path diagrams

```{r dot, eval=FALSE}
dot_ff  <- nmf.ffb.DOT(fit, model = "null",     threshold = 0.05)
dot_ffb <- nmf.ffb.DOT(fit, model = "selected", threshold = 0.05)
# plot(dot_ff); plot(dot_ffb)   # requires DiagrammeR
```

## Running the bootstrap in parallel

Both bootstraps take `cores` (or `ncores`), defaulting to
`getOption("mc.cores", 1L)` as elsewhere in the package.  With $B = 1000$ this is
worth setting:

```{r parallel, eval=FALSE}
options(mc.cores = 8)
tst <- nmf.ffb.test(fit, Y1.H1, Y2, B = 1000)
inf <- nmf.ffb.inference(fit, Y1.H1, Y2, B = 1000)
```

## Appendix: the legacy estimator

`method = "mu"` is the multiplicative-update estimator of the first version of
this work, kept for reproducibility.  It minimizes a squared error in the
structural form and cannot recover $\Theta_1$, for the reason in the Overview: it
fits the mean, and the mean does not identify feedback.

```{r mu, eval=FALSE}
fit_mu <- nmf.ffb(Y1.H1, Y2, rank = Q0, method = "mu", maxit = 500,
                  C1.L1 = 1, C2.L1 = 0.1)
```

`C1.L1` and `C2.L1` are the L1 penalties of *that* estimator and belong to it
alone.  Under `method = "fiml"` the penalty on $\Theta_1$ is the path
`C1.L1.path`, swept with the support chosen by BIC, and $\Theta_2$ is
unpenalized; passing `C1.L1` to a likelihood fit warns and does nothing.
