---
title: "Co-clustering two variable blocks with NMF-RRR (nmf.rrr)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-clustering two variable blocks with NMF-RRR (nmf.rrr)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Data-dependent chunks run only when ade4 (source of the Doubs data) is
# installed, so the vignette still builds on machines without it.
has_ade4 <- requireNamespace("ade4", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = has_ade4
)
```

## Introduction

Many studies measure **two blocks of variables on the same individuals** -- a
block of *covariates* (inputs) and a block of *responses* (outputs) -- and ask
how *groups* of covariates relate to *groups* of responses. `nmf.rrr()` answers
this by starting from the multivariate linear regression
\eqn{Y_1 \approx M Y_2} of the responses on the covariates and giving its
non-negative regression coefficient matrix a **tri-factorization**

\deqn{Y_1 \;\approx\; X_1\,\Theta\,X_2\,Y_2,\qquad X_1,X_2\ge 0,\ \Theta\ge 0,}

where \eqn{X_1} (each column summing to one) softly clusters the **response
variables**, \eqn{X_2} (each row summing to one) softly clusters the
**covariate variables**, and \eqn{\Theta} is a tested \eqn{Q\times R} matrix of
**block correspondences**. Because \eqn{M = X_1\Theta X_2} has
rank \eqn{\le \min(Q,R)}, this is the non-negative, parts-based member of the
reduced-rank regression (RRR) family -- hence *NMF-RRR* -- related to RRR as
NMF is to PCA.

`nmf.rrr()` and its `nmf.rrr.*` helpers are thin aliases for the `nmfae()`
family; the names emphasise the reduced-rank-regression reading.

This vignette reproduces the **Doubs** example (community ecology), where a
single dominant upstream--downstream gradient aligns both blocks, so the
correspondence \eqn{\Theta} is a clean *permutation*.

```{r lib}
library(nmfkc)
```

```{r no-ade4, eval = !has_ade4, echo = FALSE, results = "asis"}
cat("> **Note:** this vignette needs the `ade4` package for the Doubs data.",
    "Install it with `install.packages(\"ade4\")` to run the code below.\n")
```

## The Doubs data

The Doubs data (Verneaux, 1973) record `27` fish species and `11` environmental
variables at `30` sites along a French river -- a classic illustration of
canonical (correspondence) analysis. We take the fish abundances as the
**response** block \eqn{Y_1} and the environmental variables as the
**covariate** block \eqn{Y_2}.

Each variable is mapped to \eqn{[0,1]} by a per-variable min--max transform
(`nmfkc.normalize()`), which makes the sign-free environmental variables
non-negative, and both blocks are laid out as *variables* \eqn{\times}
*sites* (\eqn{P \times N}).

```{r data}
data(doubs, package = "ade4")

# per-variable min-max to [0,1], then transpose to (variables x sites)
nz <- function(M) t(nmfkc.normalize(as.matrix(M)))
Y1 <- nz(doubs$fish)   # responses: 27 fish species x 30 sites
Y2 <- nz(doubs$env)    # covariates: 11 environment x 30 sites
dim(Y1)
dim(Y2)
```

## Fitting NMF-RRR

Element-wise cross-validation (below) selects \eqn{Q=R=2}. Signed models and
inference benefit from several k-means restarts, so we set `nstart = 20` and a
tight tolerance.

```{r fit}
fit <- nmf.rrr(Y1, Y2, rank1 = 2, rank2 = 2,
               epsilon = 1e-8, nstart = 20, seed = 1)

# in-sample, column-centered R^2
Y1hat <- fit$X1 %*% fit$C %*% fit$X2 %*% Y2
R2 <- 1 - sum((Y1 - Y1hat)^2) / sum((Y1 - rowMeans(Y1))^2)
round(R2, 3)
```

The fit is `R2 = 0.435`, reproducing the classical longitudinal zonation of the
river.

### Response groups (fish guilds)

Each column of \eqn{X_1} is a probability vector over the fish species; the top
species per column name the guild.

```{r resp-groups}
for (q in 1:ncol(fit$X1))
  cat(sprintf("Resp%d: %s\n", q,
      paste(rownames(Y1)[order(-fit$X1[, q])[1:6]], collapse = ", ")))
```

`Resp1` is a **cold-water upstream guild** (brown trout *Satr*, *Phph*, *Neba*,
*Cogo*, grayling *Thth*) and `Resp2` a **warm-water downstream guild** (roach
*Ruru*, *Gogo*, barbel *Baba*, *Alal*).

### Covariate groups (environmental gradients)

Each row of \eqn{X_2} is a probability vector over the environmental variables.

```{r cov-groups}
for (r in 1:nrow(fit$X2))
  cat(sprintf("Cov%d: %s\n", r,
      paste(rownames(Y2)[order(-fit$X2[r, ])[1:5]], collapse = ", ")))
```

`Cov1` is a **nutrient / downstream gradient** (distance from source `dfs`,
flow `flo`, nitrate `nit`, BOD `bdo`) and `Cov2` an **oxic / upstream gradient**
(dissolved oxygen `oxy`, altitude `alt`, pH, slope `slo`).

## Choosing the two ranks

Because the attainable fit is bounded by \eqn{\min(Q,R)}, the *in-sample* fit
cannot choose the ranks; we use **element-wise cross-validation**
(`nmf.rrr.ecv()`), which holds out entries of \eqn{Y_1} and predicts them.

```{r ecv}
ecv <- nmf.rrr.ecv(Y1, Y2, rank1 = 1:2, rank2 = 1:2,
                   nfolds = 5, seed = 123)
round(ecv$sigma, 4)
```

The smallest hold-out error is at \eqn{Q=R=2}.

## Inference for the correspondence matrix \eqn{\Theta}

The entries of \eqn{\Theta} say how strongly each covariate group drives each
response group. `nmf.rrr.inference()` attaches standard errors (Fisher +
wild bootstrap) and a **one-sided boundary test** \eqn{H_0:\theta_{qr}=0}
(each \eqn{\theta_{qr}\ge 0}).

```{r inference}
inf <- nmf.rrr.inference(fit, Y1, Y2)
co  <- inf$coefficients
print(format(co[order(co$p_value), c("Basis","Covariate","Estimate","SE","z_value","p_value")],
             digits = 3))
```

\eqn{\Theta} is a **near-permutation**: the upstream guild is driven by the
oxic gradient and the downstream guild by the nutrient gradient (both
\eqn{p<0.001}), while the two off-diagonal paths are essentially zero
(\eqn{p=0.5}).

```{r theta}
round(fit$C, 3)
```

## Visualising \eqn{\Theta} and the two co-clusterings

`nmf.rrr.heatmap()` shows the response basis \eqn{X_1}, the correspondence
\eqn{\Theta}, and the covariate basis \eqn{X_2} together.

```{r heatmap, fig.width = 7.5, fig.height = 4}
nmf.rrr.heatmap(fit)
```

## Relation to other methods

Dropping non-negativity, \eqn{Y_1\approx M Y_2} at rank \eqn{\min(Q,R)} is
ordinary **reduced-rank regression** (RRR): it attains a higher in-sample fit
(\eqn{R^2\approx 0.66} on Doubs) but returns *signed* loadings and no clusters.
On these data the two share the dominant fitted direction (leading
principal-angle cosine \eqn{\approx 0.99}); they differ in the *basis* of that
subspace -- non-negative parts versus signed singular directions -- exactly as
NMF relates to PCA. An unsupervised **tri-NMF** of the association
\eqn{Y_1 Y_2^\top} recovers the *same* guilds and gradients here (the gradient
is so dominant that supervised and unsupervised co-clusterings coincide), but,
unlike NMF-RRR, cannot predict the community at a new site through
\eqn{\hat y_1 = X_1\Theta X_2\, y_2}.

When within-block and cross-block structure disagree -- e.g. under
\eqn{p>n} -- the non-negative, normalized parameterization of NMF-RRR stays
well-behaved and exposes *cross-structure* (one response group driven by
several covariate groups) that these baselines miss; see the paper for the
nutrimouse and microbiome--metabolome examples.

## References

- Satoh, K. & Tokuda, Y. Co-clustering of Response and Covariate Variables by
  Tri-Factorizing Their Non-negative Regression Coefficient Matrix
  (manuscript).
- Ding, C., Li, T., Peng, W. & Park, H. (2006). Orthogonal nonnegative matrix
  tri-factorizations for clustering. *KDD*.
- Verneaux, J. (1973). *Cours d'eau de Franche-Comté*. PhD thesis. (Doubs data,
  R package `ade4`.)
```
