---
title: "NMF-COX: Interpretable Non-Proportional Hazards with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-COX: Interpretable Non-Proportional Hazards with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{survival}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Introduction

The Cox proportional-hazards model assumes that each covariate acts with a
time-constant log hazard ratio. When that assumption fails --- the effect of a
covariate strengthens or fades over follow-up --- the usual coefficient is hard
to interpret. **NMF-COX** models the non-proportional part explicitly with a
low-rank, non-negative time-varying offset and keeps a clean, time-constant
effect for the covariates of interest.

The hazard is
$$\lambda_n(t) = h_0(t)\exp\!\big(\boldsymbol z_n^\top\gamma + \boldsymbol a_n^\top\beta(t)\big),
\qquad \beta(t) = C^\top \boldsymbol x(t),\quad \boldsymbol x(t) \ge 0.$$

- $\boldsymbol z$ are the **proportional-hazards covariates of interest**; their
  effect $\gamma$ is time-constant and is the *estimand*.
- $\boldsymbol a$ are the **non-proportional covariates**; they receive a
  time-varying coefficient $\beta(t)$.
- $X$ ($P\times Q$) is a **non-negative time basis** over the $P$ event times
  (columns normalised to sum to one, i.e. composition profiles), and $C$
  ($Q\times R$; $\Theta$ in the paper) is a **sign-free parameter matrix** ---
  the combination is a semi-NMF representation, estimated directly on the Cox
  partial likelihood (Breslow ties).

Because the offset is linear in $\boldsymbol a$ at each event time, $\gamma$ is
separated from the time-varying part and can be estimated by a cross-fit
(Neyman-orthogonal) procedure. The non-negative, sum-to-one basis makes each
$\beta_r(t)$ readable as a *time-phase profile*, unlike a sign-free reduced-rank
fit whose components oscillate and are hard to interpret.

Following the rest of the package, `nmf.cox()` performs **optimization only**;
`nmf.cox.inference()` adds standard errors and tests (mirroring the
`nmfkc()` / `nmfkc.inference()` split).

---

## 1. Data: the `veteran` lung-cancer trial

We use the `veteran` data from the `survival` package (137 patients, 128
deaths). We treat treatment (`trt`) and `age` as the covariates of interest
$\boldsymbol z$, and the Karnofsky performance score (`karno`) and cell type
(`celltype`) as the potentially non-proportional covariates $\boldsymbol a$
--- Karnofsky score is the classic textbook example of a fading effect.

```{r data}
library(nmfkc)
library(survival)

data(veteran)
veteran$y <- Surv(veteran$time, veteran$status)
```

---

## 2. Fitting the model (optimization only)

`A` accepts a one-sided formula for the non-proportional covariates. We use a
rank-2 time basis and a roughness penalty `X.L2.smooth` on $\beta(t)$ (chosen
in Section 5).

```{r fit}
fit <- nmf.cox(y ~ trt + age, data = veteran, A = ~ karno + celltype,
               rank = 2, X.L2.smooth = 1000, verbose = FALSE)
fit
```

The printed proportional-hazards coefficients $\gamma$ (for `trt`, `age`) come
from a Cox fit that carries the estimated time-varying offset for
`karno`/`celltype`. The object follows the house style:

```{r fit-fields}
fit$dims
round(fit$gamma, 4)      # gamma: time-constant effects of z (trt, age)
dim(fit$C)               # C (= Theta): Q x R sign-free parameter matrix
```

---

## 3. Inference: `nmf.cox.inference()`

To obtain standard errors of $\beta(t)$, the coefficients table for $C$, and the
Wald tests, pass the fit and the same model specification to
`nmf.cox.inference()`. It returns the object with the inference fields added, so
`summary()` now prints the full report.

```{r inference}
fit <- nmf.cox.inference(fit, y ~ trt + age, data = veteran, A = ~ karno + celltype)
summary(fit)
```

The **Wald test** is of $H_0:\beta_r(t)=0$ for each non-proportional covariate,
*conditional on the estimated basis* $\hat X$: a significant result means the
covariate has a genuinely time-varying effect. Here the Karnofsky score has a
strong non-proportional effect. The **coefficients table for $C$** uses the
house-style `Basis` / `Covariate` columns:

```{r coef-table}
head(fit$coefficients, 6)
```

For inference that also accounts for having *selected* the basis on the data,
use the cross-fitted tests in Sections 6--7.

---

## 4. Reading the time-varying effect $\beta(t)$

`fitted()` returns $\beta(t) = XC$ at the event times (rows are event times,
columns are covariates), with pointwise standard errors in `fit$se.beta.t`.

```{r beta-plot, fig.height=5, fig.alt="Estimated time-varying log hazard ratio of the Karnofsky score with a 95% confidence band; strongly negative early and attenuating over follow-up."}
et  <- fit$event.times
B   <- fitted(fit)                 # beta(t): P x R
SE  <- fit$se.beta.t
cov <- "karno"

est <- B[, cov]; lo <- est - 1.96 * SE[, cov]; hi <- est + 1.96 * SE[, cov]
plot(et, est, type = "l", lwd = 2, col = "steelblue",
     ylim = range(lo, hi, 0),
     xlab = "Time (days)", ylab = expression(beta[karno](t)),
     main = "Time-varying log hazard ratio of Karnofsky score (per SD)")
polygon(c(et, rev(et)), c(lo, rev(hi)),
        col = adjustcolor("steelblue", 0.2), border = NA)
lines(et, est, lwd = 2, col = "steelblue")
abline(h = 0, lty = 2, col = "grey40")
```

The Karnofsky effect is strongly **protective (negative log hazard ratio) early**
and **attenuates over follow-up** --- the textbook non-proportional pattern that
a single Cox coefficient would average away.

The non-negative basis itself is directly interpretable as time-phase profiles
(each column of `X.prob` sums to one over event times):

```{r basis-plot, fig.alt="The two non-negative time-basis columns as composition-probability profiles over event times, each concentrating on a different time phase."}
matplot(et, fit$X.prob, type = "l", lwd = 2, lty = 1,
        col = c("darkorange", "purple"),
        xlab = "Time (days)", ylab = "Composition probability",
        main = "Non-negative time basis (phase profiles)")
legend("topright", colnames(fit$X.prob), col = c("darkorange", "purple"),
       lwd = 2, bty = "n")
```

---

## 5. Choosing the rank and smoothing: `nmf.cox.cv()`

`nmf.cox.cv()` sweeps vectors of `rank` and `X.L2.smooth`, by the
cross-validated partial likelihood (CVPL) or a partial-likelihood AIC/BIC. It
returns `rank.best` / `X.L2.smooth.best` (and, for CVPL, the parsimonious
1-SE variants). The AIC criterion needs a single fit per grid point and is a
fast proxy.

```{r cv}
cv <- nmf.cox.cv(y ~ trt + age, data = veteran, A = ~ karno + celltype,
                 rank = 2, X.L2.smooth = c(100, 300, 1000, 3000),
                 criterion = "aic", verbose = FALSE)
cv$X.L2.smooth.best
round(cv$aic, 1)
```

---

## 6. The estimand $\gamma$: cross-fit inference with `nmf.cox.cf()`

The clean, time-constant effect of the covariates of interest is estimated by a
two-stage cross-fit: for each fold the time-varying nuisance $\beta^{(-k)}(t)$
is learned *without* that fold, its out-of-fold offset is frozen, and $\gamma$
is estimated on the full sample with cluster-robust standard errors.

```{r cf}
cf <- nmf.cox.cf(y ~ trt + age, data = veteran, A = ~ karno + celltype,
                 rank = 2, X.L2.smooth = 1000, nfolds = 5, seed = 1)
cf
```

It is instructive to compare with a **standard Cox model that ignores the
non-proportionality** of `karno`/`celltype`:

```{r naive}
naive <- coxph(Surv(time, status) ~ trt + age, data = veteran)
round(cbind(`NMF-COX (cross-fit)` = cf$gamma,
            `standard Cox`        = coef(naive)), 4)
```

Once the time-varying effect of the strong prognostic covariates is modelled,
the estimand $\gamma$ for the covariates of interest can differ appreciably from
the naive adjustment.

---

## 7. Cross-fitted test of a time-varying effect: `nmf.cox.phtest()`

Section 3's Wald test conditions on the basis learned from the same data.
`nmf.cox.phtest()` instead learns the basis **out of fold**, so the test of
$H_0:\beta_r(t)=0$ is not invalidated by basis selection.

```{r phtest}
ph <- nmf.cox.phtest(y ~ trt + age, data = veteran, A = ~ karno + celltype,
                     rank = 2, X.L2.smooth = 1000, nfolds = 5, seed = 1)
ph
```

---

## Summary

| Step | Function | Purpose |
|:---|:---|:---|
| 1 | `nmf.cox()` | Fit the model (optimization only) |
| 2 | `nmf.cox.inference()` | SEs of $\beta(t)$, coefficients table for $C$, given-basis Wald tests |
| 3 | `nmf.cox.cv()` | Select `rank` and `X.L2.smooth` (CVPL / AIC / BIC) |
| 4 | `nmf.cox.cf()` | Cross-fit estimand $\gamma$ for the covariates of interest |
| 5 | `nmf.cox.phtest()` | Cross-fitted test of a time-varying effect |

**When to use NMF-COX:** survival data where some covariates violate
proportional hazards in an interpretable, gradually-changing way, and you still
want a clean time-constant effect for the covariates of interest.

For methodological details see Satoh, K. (2026), *NMF-COX: Interpretable
modelling of non-proportional hazards and cross-fit inference for the
proportional effect*.
