---
title: "NMF-FFB with nmfkc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMF-FFB with nmfkc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{lavaan}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette shows how to fit **NMF-FFB** (non-negative matrix
factorization with feedforward and feedback structure) with **nmfkc**,
and how to decide whether the feedback is supported by the data.

NMF with covariates (the *feedforward* model, NMF-FF) approximates a
non-negative endogenous data matrix $Y_1$ ($P_1 \times N$) by a
column-stochastic basis $X$ and covariate-driven scores,

$$
Y_1 \approx X\,\Theta_2 Y_2 ,
$$

where $Y_2$ ($P_2 \times N$) holds the exogenous variables. NMF-FFB adds a
non-negative **feedback** term through which the endogenous variables act
back on the latent factors,

$$
Y_1 \approx X(\Theta_1 Y_1 + \Theta_2 Y_2).
$$

Two facts shape the estimator (see `?nmf.ffb` and the references):

* The equilibrium map $M=(I-X\Theta_1)^{-1}X\Theta_2$ of any stable NMF-FFB
  model can be reproduced exactly by an NMF-FF model with the same basis
  (reduced-form equivalence). Predictive fit therefore cannot tell
  feedback from its absence.
* Feedback is identified through disturbances that propagate around the
  loop and through an **exclusion restriction**: an outcome may not feed
  back into the factor on which it loads (no self-loops).

`nmf.ffb()` therefore works in two stages: the basis $X$ is estimated by
`nmfkc()` (Stage 1), and, conditional on $X$, a Gaussian working model

$$
Y_1 = XB + \mathcal{E},\qquad B=\Theta_1Y_1+\Theta_2Y_2+U,\qquad
U\sim N(0,\Phi),\quad \mathcal{E}\sim N(0,\Psi)
$$

is fitted by full-information maximum likelihood (Stage 2). The
feedforward model ($\Theta_1=0$) is the nested null model, a sparse
$\Theta_1$ is chosen by an L1 path with BIC, and the likelihood-ratio
statistic is calibrated by a parametric bootstrap in
`nmf.ffb.inference()`.

---

## 1. Data: Holzinger--Swineford tests

```{r}
library(lavaan)
library(nmfkc)
data(HolzingerSwineford1939)

d <- HolzingerSwineford1939
d <- d[complete.cases(d), ]

## exogenous variables: reversed age, sex, school
d$age.rev   <- -(d$ageyr + d$agemo / 12)
d$sex.2     <- ifelse(d$sex == 2, 1, 0)
d$school.GW <- ifelse(d$school == "Grant-White", 1, 0)
d[, c("id", "sex", "ageyr", "agemo", "school", "grade")] <- NULL

## min-max scaling to [0, 1] (all variables non-negative)
d <- nmfkc.normalize(d)

exogenous_vars  <- c("age.rev", "sex.2", "school.GW")
endogenous_vars <- setdiff(colnames(d), exogenous_vars)
Y1 <- t(d[, endogenous_vars])   # 9 tests x 300 pupils
Y2 <- t(d[, exogenous_vars])    # 3 covariates x 300 pupils
```

## 2. Choosing the number of factors

The number of factors $Q$ is a property of the feedforward stage, so it is
chosen by element-wise cross-validation of the feedforward fit
(`nmf.ffb.cv()` delegates to `nmfkc.ecv()` for `method = "fiml"`;
column-wise cross-validation of the equilibrium prediction cannot select
the feedback penalty because of the reduced-form equivalence).

```{r, eval = FALSE}
cv <- nmf.ffb.cv(Y1, Y2, rank = 2:4, method = "fiml", X.L2.ortho = 100)
cv$objfunc      # held-out reconstruction error by rank
```

For these data the canonical three-factor structure (visual, textual,
speed) is used, $Q = 3$.

## 3. Fitting NMF-FFB

`nmf.ffb()` estimates the basis with `nmfkc()` (NNDSVDar initialization,
orthogonality penalty `X.L2.ortho`) and then fits the working model by
FIML. By default the exclusion mask is `"block"` (an outcome may feed
back only into factors other than its dominant one), $\Phi$ is a full
covariance matrix, and the sparse feedback structure is selected by BIC.

```{r}
Q0  <- 3
res <- nmf.ffb(Y1, Y2, rank = Q0, X.L2.ortho = 100)   # method = "fiml" (default)
summary(res)
```

The summary reports the log-likelihoods of the null (feedforward), the
unpenalized feedback model and the BIC-selected model, the two
likelihood-ratio statistics, the number of retained feedback entries and
the spectral radius $\rho(X\Theta_1)$ of the selected feedback operator.

```{r}
round(res$X, 2)          # basis (column-stochastic): which test loads on which factor
round(res$C1, 2)         # selected feedback Theta1 (Q x P1): outcome -> factor
round(res$C2, 2)         # exogenous coefficients Theta2 (Q x P2)
res$LR                   # LR statistics vs. the feedforward null (full / selected)
res$BIC                  # BIC of null / full / selected
res$XC1.radius           # rho(X Theta1) of the selected model
```

The BIC path shows how the number of retained feedback entries and the
log-likelihood change along the L1 penalty. Because the penalized problem
is non-convex, every penalty is fitted from several starting points
(`starts`); `path` has one row per (penalty, start) and flags the rows
that re-propose an already known support, and `candidates` lists every
distinct support with its BIC. On these data the single start `"full"`
would have stopped at a one-path model; the six-path model with the
smallest BIC is found by the `"soft"` and `"null"` starts.

```{r}
res$path[!res$path$duplicate, ]   # first proposal of each distinct support
res$candidates                    # all distinct supports; BIC is minimized over these
plot(res)                # BIC versus log(lambda1); the selected point is circled
```

## 4. Is the feedback supported? Parametric bootstrap

Because $\Theta_1 \ge 0$ puts the null on the boundary and the support is
selected from the data, the LR statistics do not follow a chi-square
distribution. `nmf.ffb.inference()` simulates from the fitted feedforward
model, re-runs the whole selection pipeline on each replicate, and
returns bootstrap $p$-values together with the *null false-selection
rate*, the probability that BIC retains at least one spurious entry when
the feedforward model is true. A second bootstrap from the selected model
gives interval estimates of the retained coefficients.

```{r}
inf <- nmf.ffb.inference(res, Y1, Y2, B = 20, seed = 1)   # use B = 1000 in practice
inf$LR.p.boot            # bootstrap p-values for LR(full) and LR(selected)
inf$prob.select.null     # null false-selection rate of BIC
head(inf$coefficients)   # retained coefficients with bootstrap intervals
```

With a small `B` the $p$-values are coarse; the published analysis uses
`B = 1000` (about one minute on a few cores because each replicate refits
the whole path). Pass `cores = ` to parallelize.

## 5. Path diagrams

`nmf.ffb.DOT()` draws the feedforward model (`model = "null"`) and the
BIC-selected model (`model = "selected"`) on the same node layout, so the
effect of the feedback is visible as the difference between the two
diagrams; feedback edges are dashed.

```{r}
dot_ff  <- nmf.ffb.DOT(res, model = "null",     threshold = 0.05)
dot_ffb <- nmf.ffb.DOT(res, model = "selected", threshold = 0.05)
# plot(dot_ff); plot(dot_ffb)   # requires DiagrammeR
```

## 6. Reading the result

In the Holzinger--Swineford data the three factors reproduce the
visual/textual/speed grouping. The selected feedback links the speeded
tests to the visual factor and the visual tests to the speed factor, a
cross-factor cycle with $\rho(X\Theta_1)\approx 0.6$, and the bootstrap
places the observed LR far outside the null distribution. Two cautions
apply in general: the feedback coefficients are conditional structural
coefficients, not causal effects, and a cross-factor cycle can absorb
part of what a factor correlation would otherwise explain, so the change
in $\Phi$ between the null and the selected fit should be inspected
(`res$null$Phi` versus `res$Phi`).

## Appendix: the legacy estimator

The multiplicative-update estimator of the first version of the method
(structural least squares with the observed $Y_1$ on the right-hand side)
is kept as `method = "mu"` for reproducibility. It does not identify
$\Theta_1$---its estimate of $\rho(X\Theta_1)$ is governed by the L1
penalty and the sample size rather than by the true feedback---and it
will be deprecated in a future release.

```{r, eval = FALSE}
res_mu <- nmf.ffb(Y1, Y2, rank = Q0, X.L2.ortho = 100,
                  C1.L1 = 10, C2.L1 = 0.6, method = "mu")
```
