---
title: "Co-clustering two variable blocks with NMF-RRR (nmf.rrr)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Co-clustering two variable blocks with NMF-RRR (nmf.rrr)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Data-dependent chunks run only when ade4 (source of the Doubs data) is
# installed, so the vignette still builds on machines without it.
has_ade4 <- requireNamespace("ade4", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = has_ade4
)
```

## Introduction

Many studies measure **two blocks of variables on the same individuals** -- a
block of *covariates* (inputs) and a block of *responses* (outputs) -- and ask
how *groups* of covariates relate to *groups* of responses. `nmf.rrr()` answers
this by starting from the multivariate linear regression
\eqn{Y_1 \approx M Y_2} of the responses on the covariates and giving its
non-negative regression coefficient matrix a **tri-factorization**

\deqn{Y_1 \;\approx\; X_1\,\Theta\,X_2\,Y_2,\qquad X_1,X_2\ge 0,\ \Theta\ge 0,}

where \eqn{X_1} (each column summing to one) softly clusters the **response
variables**, \eqn{X_2} (each row summing to one) softly clusters the
**covariate variables**, and \eqn{\Theta} is a tested \eqn{Q\times R} matrix of
**block correspondences**. Because \eqn{M = X_1\Theta X_2} has
rank \eqn{\le \min(Q,R)}, this is the non-negative, parts-based member of the
reduced-rank regression (RRR) family -- hence *NMF-RRR* -- related to RRR as
NMF is to PCA.

`nmf.rrr()` and its `nmf.rrr.*` helpers are the canonical interface; the
names emphasise the reduced-rank-regression reading. The former `nmfae()`
family names are retained as deprecated aliases for backward compatibility.

This vignette reproduces the **Doubs** example (community ecology), where a
single dominant upstream--downstream gradient aligns both blocks, so the
correspondence \eqn{\Theta} is a clean *permutation*.

```{r lib}
library(nmfkc)
```

```{r no-ade4, eval = !has_ade4, echo = FALSE, results = "asis"}
cat("> **Note:** this vignette needs the `ade4` package for the Doubs data.",
    "Install it with `install.packages(\"ade4\")` to run the code below.\n")
```

## The Doubs data

The Doubs data (Verneaux, 1973) record `27` fish species and `11` environmental
variables at `30` sites along a French river -- a classic illustration of
canonical (correspondence) analysis. We take the fish abundances as the
**response** block \eqn{Y_1} and the environmental variables as the
**covariate** block \eqn{Y_2}.

Each variable is mapped to \eqn{[0,1]} by a per-variable min--max transform
(`nmfkc.normalize()`), which makes the sign-free environmental variables
non-negative, and both blocks are laid out as *variables* \eqn{\times}
*sites* (\eqn{P \times N}).

```{r data}
data(doubs, package = "ade4")

# per-variable min-max to [0,1], then transpose to (variables x sites)
nz <- function(M) t(nmfkc.normalize(as.matrix(M)))
Y1 <- nz(doubs$fish)   # responses: 27 fish species x 30 sites
Y2 <- nz(doubs$env)    # covariates: 11 environment x 30 sites
dim(Y1)
dim(Y2)
```

## Fitting NMF-RRR

Element-wise cross-validation (below) selects \eqn{Q=R=2}. Signed models and
inference benefit from several k-means restarts, so we set `nstart = 20` and a
tight tolerance.

```{r fit}
fit <- nmf.rrr(Y1, Y2, rank1 = 2, rank2 = 2,
               epsilon = 1e-8, nstart = 20, seed = 1)

# in-sample, column-centered R^2
Y1hat <- fit$X1 %*% fit$C %*% fit$X2 %*% Y2
R2 <- 1 - sum((Y1 - Y1hat)^2) / sum((Y1 - rowMeans(Y1))^2)
round(R2, 3)
```

The fit is `R2 = 0.435`, reproducing the classical longitudinal zonation of the
river.

### Response groups (fish guilds)

Each column of \eqn{X_1} is a probability vector over the fish species; the top
species per column name the guild.

```{r resp-groups}
for (q in 1:ncol(fit$X1))
  cat(sprintf("Resp%d: %s\n", q,
      paste(rownames(Y1)[order(-fit$X1[, q])[1:6]], collapse = ", ")))
```

`Resp1` is a **cold-water upstream guild** (brown trout *Satr*, *Phph*, *Neba*,
*Cogo*, grayling *Thth*) and `Resp2` a **warm-water downstream guild** (roach
*Ruru*, *Gogo*, barbel *Baba*, *Alal*).

### Covariate groups (environmental gradients)

Each row of \eqn{X_2} is a probability vector over the environmental variables.

```{r cov-groups}
for (r in 1:nrow(fit$X2))
  cat(sprintf("Cov%d: %s\n", r,
      paste(rownames(Y2)[order(-fit$X2[r, ])[1:5]], collapse = ", ")))
```

`Cov1` is a **nutrient / downstream gradient** (distance from source `dfs`,
flow `flo`, nitrate `nit`, BOD `bdo`) and `Cov2` an **oxic / upstream gradient**
(dissolved oxygen `oxy`, altitude `alt`, pH, slope `slo`).

## Choosing the two ranks

Because the attainable fit is bounded by \eqn{\min(Q,R)}, the *in-sample* fit
cannot choose the ranks; we use **element-wise cross-validation**
(`nmf.rrr.ecv()`), which holds out entries of \eqn{Y_1} and predicts them.

```{r ecv}
ecv <- nmf.rrr.ecv(Y1, Y2, rank1 = 1:2, rank2 = 1:2,
                   nfolds = 5, seed = 123)
round(ecv$sigma, 4)
```

The smallest hold-out error is at \eqn{Q=R=2}.

## Inference for the correspondence matrix \eqn{\Theta}

The entries of \eqn{\Theta} say how strongly each covariate group drives each
response group. `nmf.rrr.inference()` attaches standard errors (Fisher +
wild bootstrap) and a **one-sided boundary test** \eqn{H_0:\theta_{qr}=0}
(each \eqn{\theta_{qr}\ge 0}).

```{r inference}
inf <- nmf.rrr.inference(fit, Y1, Y2)
co  <- inf$coefficients
print(format(co[order(co$p_value), c("Basis","Covariate","Estimate","SE","z_value","p_value")],
             digits = 3))
```

\eqn{\Theta} is a **near-permutation**: the upstream guild is driven by the
oxic gradient and the downstream guild by the nutrient gradient (both
\eqn{p<0.001}), while the two off-diagonal paths are essentially zero
(\eqn{p=0.5}).

```{r theta}
round(fit$C, 3)
```

## Visualising \eqn{\Theta} and the two co-clusterings

`nmf.rrr.heatmap()` shows the response basis \eqn{X_1}, the correspondence
\eqn{\Theta}, and the covariate basis \eqn{X_2} together.

```{r heatmap, fig.width = 7.5, fig.height = 4}
nmf.rrr.heatmap(fit)
```

## Clustering the individuals as well as the variables

The same fit clusters the **sites**. The two score matrices

\deqn{B_2 = X_2 Y_2 \quad (R\times N), \qquad B_1 = \Theta B_2 \quad (Q\times N)}

give each site a profile over the covariate groups and over the response
groups; column-normalising them yields soft memberships (`B2.prob`,
`B1.prob`), and the arg-max gives hard labels (`B2.cluster`, `B1.cluster`).
So a single model produces four groupings: two of variables (\eqn{X_1},
\eqn{X_2}) and two of individuals (\eqn{B_1}, \eqn{B_2}).

```{r ind-algebra}
max(abs(fit$B1 - fit$C %*% fit$X2 %*% Y2))   # B1 = Theta B2, to machine precision
```

The Doubs sites are numbered from the source (1) downstream to (30), **and
that order is never given to the model**. A labelling that comes out
contiguous along the river is therefore a genuine recovery of the
longitudinal gradient, not something the fit was told.

```{r ind-clusters}
runs <- function(cl, tag) {
  cl <- as.vector(cl); r <- rle(cl)
  cat(sprintf("%-4s %s\n", tag, paste(cl, collapse = " ")))
  cat(sprintf("%-4s %d run(s): %s\n", "", length(r$lengths),
              paste(sprintf("%s x%d", r$values, r$lengths), collapse = " | ")))
  if (length(r$lengths) == 2)
    cat(sprintf("%-4s one break, between site %d and %d\n", "",
                r$lengths[1], r$lengths[1] + 1))
  cat("\n")
}
runs(fit$B1.cluster, "B1")   # response side: fish community
runs(fit$B2.cluster, "B2")   # covariate side: environment
```

Each side splits the river once, but **not at the same place**: the fish
community changes at site 10|11, the environment at 20|21. Sites 11--20 are
already downstream-type in their fauna while their measured environment is
still upstream-type -- the fish respond before the environmental variables
recorded here cross over.

The soft memberships show how sharp each transition is.

```{r ind-soft}
round(fit$B1.prob[, 8:13], 2)    # response side, around the 10|11 break
round(fit$B2.prob[, 18:23], 2)   # covariate side, around the 20|21 break
```

### Mapping the clusters onto the river

`doubs$xy` holds schematic map coordinates for the 30 sites, so the two
labellings can be drawn on the river itself.

```{r ind-map, fig.width = 8, fig.height = 4.2}
xy <- doubs$xy
panel <- function(cl, upstream.level, main, brk) {
  fill <- ifelse(as.vector(cl) == upstream.level, "white", "black")
  plot(xy, type = "n", asp = 1, xlab = "x (km)", ylab = "y (km)", main = main)
  lines(xy, col = "grey60", lwd = 2)                    # river course
  points(xy, pch = 21, bg = fill, col = "black", cex = 1.6)
  text(xy, labels = 1:30, pos = 3, offset = 0.4, cex = 0.55)
  mid <- (xy[brk, ] + xy[brk + 1, ]) / 2                # mark the break
  points(mid, pch = 4, cex = 2, lwd = 2.5)
}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
panel(fit$B1.cluster, 1, "B1: fish community", 10)
panel(fit$B2.cluster, 2, "B2: environment", 20)
par(op)
```

Open circles are the upstream type, filled the downstream type, and the cross
marks the break. Both transitions are single and contiguous, and the offset
between them is the stretch where the two blocks disagree.

Whether supervision matters here is worth checking: `k-means` on either raw
block, blind to the other block, is the natural baseline. Cluster labels are
arbitrary, so agreement is measured in a label-switch-invariant way.

```{r ind-baseline}
set.seed(1)
km1 <- kmeans(t(Y1), 2, nstart = 50)$cluster
km2 <- kmeans(t(Y2), 2, nstart = 50)$cluster
agree <- function(a, b) max(mean(a == b), mean(a != b))
c(B1.vs.kmY1 = agree(km1, as.vector(fit$B1.cluster)),
  B2.vs.kmY2 = agree(km2, as.vector(fit$B2.cluster)))
runs(km1, "km1")   # k-means on the fish block alone
runs(km2, "km2")   # k-means on the environment block alone
```

The covariate side is close (`0.87`): the environmental gradient is strong
enough that an unsupervised split of \eqn{Y_2} nearly reproduces \eqn{B_2}.
The response side is not (`0.60`). Left to itself, `k-means` on the fish block
does not even return a contiguous stretch of river -- it breaks late and then
scatters a few sites back -- because unsupervised clustering weights whatever
variance dominates \eqn{Y_1}, gradient or not.

\eqn{B_1} is contiguous because it is not computed from \eqn{Y_1} at all: it is
\eqn{\Theta B_2}, the environment read through the correspondence matrix. That
is what supervision buys here, and it is also why the two site clusterings can
be compared at all -- they are tied to each other through \eqn{\Theta}, so the
10\|11 versus 20\|21 offset means something. The same model says which fish
guild goes with which environmental gradient, and where along the river each
one turns over.

## Relation to other methods

Dropping non-negativity, \eqn{Y_1\approx M Y_2} at rank \eqn{\min(Q,R)} is
ordinary **reduced-rank regression** (RRR): it attains a higher in-sample fit
(\eqn{R^2\approx 0.66} on Doubs) but returns *signed* loadings and no clusters.
On these data the two share the dominant fitted direction (leading
principal-angle cosine \eqn{\approx 0.99}); they differ in the *basis* of that
subspace -- non-negative parts versus signed singular directions -- exactly as
NMF relates to PCA. An unsupervised **tri-NMF** of the association
\eqn{Y_1 Y_2^\top} recovers the *same* guilds and gradients here (the gradient
is so dominant that supervised and unsupervised co-clusterings coincide), but,
unlike NMF-RRR, cannot predict the community at a new site through
\eqn{\hat y_1 = X_1\Theta X_2\, y_2}.

When within-block and cross-block structure disagree -- e.g. under
\eqn{p>n} -- the non-negative, normalized parameterization of NMF-RRR stays
well-behaved and exposes *cross-structure* (one response group driven by
several covariate groups) that these baselines miss; see the paper for the
nutrimouse and microbiome--metabolome examples.

## References

- Satoh, K. & Tokuda, Y. Co-clustering of Response and Covariate Variables by
  Tri-Factorizing Their Non-negative Regression Coefficient Matrix
  (manuscript).
- Ding, C., Li, T., Peng, W. & Park, H. (2006). Orthogonal nonnegative matrix
  tri-factorizations for clustering. *KDD*.
- Verneaux, J. (1973). *Cours d'eau de Franche-Comté*. PhD thesis. (Doubs data,
  R package `ade4`.)
```
