---
title: "Time Series Analysis with NMF-VAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Analysis with NMF-VAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{vars}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to apply the `nmfkc` package to time series data using the **NMF-VAR (Non-negative Matrix Factorization - Vector Autoregression)** framework. This approach models the coefficient matrix of NMF as a VAR process ($B \approx C A_{lag}$), allowing for both dimensionality reduction and temporal modeling.

Key functions in `nmfkc` such as `nmfkc.ar` and `nmfkc.ar.degree.cv` directly support R's native **`ts` objects**, preserving time series properties automatically. Once a model is fitted, `nmfkc.ar.stationarity()` checks dynamic stability and reports the $Q \times Q$ **latent transition matrices** $G_d = \Theta_d X$, which summarise the dynamics in the low-dimensional latent space rather than in the $P \times P$ observed space.

We will cover two scenarios:

1.  **Univariate AR:** Forecasting airline passengers (`AirPassengers`).
2.  **Multivariate VAR:** Analyzing macroeconomic indicators (`Canada`).

First, let's load the necessary packages.

```{r load-packages}
library(nmfkc)
library(vars) # For Canada dataset
```

-----

## Example 1: Univariate Autoregression with AirPassengers

The `AirPassengers` dataset contains monthly international airline passenger numbers. We will model this univariate time series using an autoregressive (AR) model.

### 1\. Data Preparation

We simply log-transform the `ts` object to stabilize variance. We do **not** need to manually convert it to a matrix or create time vectors; `nmfkc` functions handle `ts` objects directly.

```{r air-data-prep}
# Load and transform the ts object
d_air <- AirPassengers
d_air_log <- log10(d_air) # Still a ts object
```

### 2\. Model Selection (Lag Order)

We use **Cross-Validation** to find the optimal degree (lag order). By passing the `ts` object `d_air_log`, the function automatically handles the time dimension.

```{r air-degree-cv}
# Evaluate lag orders from 1 to 14
# Note: ts objects are automatically transposed to (Variables x Time) internally
cv_res <- nmfkc.ar.degree.cv(d_air_log, rank = 1, degree = 1:14, epsilon=1e-6, maxit=500000)

# Check the optimal degree
cv_res$degree

# For this example, we will proceed with D=12 (capturing monthly seasonality)
D <- 12
```

### 3\. Model Fitting

We construct the observation matrix `Y` and covariate matrix `A` (lagged Y) using `nmfkc.ar()`. The returned matrices `Y` and `A` will have time-based column names derived from the `ts` object.

```{r air-model-fit}
# Create matrices for the AR(12) model
a_air <- nmfkc.ar(d_air_log, degree = D, intercept = TRUE)

# Fit the NMF-AR model (Rank=1 for univariate)
res_air <- nmfkc(Y = a_air$Y, A = a_air$A, rank = 1, epsilon = 1e-6, maxit=500000)

# Check goodness of fit
res_air$r.squared

# Check for stationarity (spectral radius < 1)
st_air <- nmfkc.ar.stationarity(res_air)
c(rho = st_air$spectral.radius, stationary = st_air$stationary)
```

With `rank = 1` the latent VAR is one-dimensional, so the latent view is at its
simplest: the twelve $1\times1$ matrices $G_d$ are just the AR coefficients of
the single latent series, and `colsum.max` alone certifies stationarity.

```{r air-latent}
# st_air was computed above
c(rho = st_air$spectral.radius, colsum.max = st_air$colsum.max)
unlist(st_air$G)[1:3]   # G_1, G_2, G_3 of the latent series
```

### 4\. Forecasting

We can forecast future values using the fitted model. `nmfkc.ar.predict` uses the time properties stored in the model to generate the correct future time sequence.

```{r air-forecast}
# Forecast next 2 years (24 months)
h <- 24
pred_res <- nmfkc.ar.predict(x = res_air, Y = a_air$Y, n.ahead = h)

# Convert predictions back to original scale
pred_val <- 10^as.vector(pred_res$pred)
pred_time <- pred_res$time # Future time points generated by the function

# --- Plotting ---
# Setup plot range
xlim_range <- range(c(time(d_air), pred_time))
ylim_range <- range(c(d_air, pred_val))

# 1. Observed data (Black)
plot(d_air, type = "l", col = "black", 
     xlim = xlim_range, ylim = ylim_range, lwd = 1,
     xlab = "Year", ylab = "Air Passengers", main = "NMF-VAR Forecast (h=24)")

# 2. Fitted values during training (Red)
# a_air$Y has column names as time strings; we parse them for plotting
fitted_time <- as.numeric(colnames(res_air$XB))
lines(fitted_time, 10^as.vector(res_air$XB), col = "red", lwd = 2)

# 3. Forecast (Blue)
# Connect the last observed point to the first forecast for a continuous line
last_t <- tail(as.numeric(time(d_air)), 1)
last_y <- tail(as.vector(d_air), 1)
lines(c(last_t, pred_time), c(last_y, pred_val), col = "blue", lwd = 2, lty = 2)

# Add legend
legend("topleft", legend = c("Observed", "Fitted", "Forecast"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = 2)
```

-----

## Example 2: Vector Autoregression with Canada Dataset

The `Canada` dataset contains four quarterly macroeconomic variables. We'll use a multivariate NMF-VAR model to analyze the relationship between these variables.

### 1\. Data Preparation

We take the first difference to achieve stationarity, normalize the data to $[0, 1]$, and transpose it to the `Variables x Time` format required by NMF.

```{r canada-data-prep}
# Load, difference, and normalize
d0_canada <- Canada
dd_canada <- apply(d0_canada, 2, diff) # Returns a matrix (Time x Vars)
dn_canada <- nmfkc.normalize(dd_canada)

# Transpose to (Variables x Time) for NMF
Y0_canada <- t(dn_canada)

# Create matrices for VAR(1)
a_canada <- nmfkc.ar(Y0_canada, degree = 1, intercept = TRUE)
```

### 2\. Model Fitting

We fit a model with `rank = 2` to identify two latent economic conditions driving the four variables.

```{r canada-model-fit}
# Fit the NMF-VAR model
# epsilon = 1e-8: this fit feeds the inference in Section 5, where coefficients
# sitting on the non-negativity boundary must actually have reached it.
res_canada <- nmfkc(Y = a_canada$Y, A = a_canada$A, rank = 2, prefix = "Condition", epsilon = 1e-8)

# R-squared and Stationarity
res_canada$r.squared
st_canada <- nmfkc.ar.stationarity(res_canada)
c(rho = st_canada$spectral.radius, stationary = st_canada$stationary)
```

### 3\. Latent Transition Matrices

The fitted VAR acts on the four observed variables through the $P \times P$
matrices $\Xi_d = X\Theta_d$. Those are awkward to read, and they grow with the
number of series. But the model implies something much smaller: because
$\boldsymbol y_t = X\boldsymbol b_t$, the **coefficient** vector follows a
non-negative VAR of its own,

$$\boldsymbol b_t=\sum_{d=1}^{D}G_d\,\boldsymbol b_{t-d}+\boldsymbol\theta,
  \qquad G_d=\Theta_d X\ \ (Q\times Q).$$

`nmfkc.ar.stationarity()` returns these **latent transition matrices**. Here that
means a single $2\times2$ matrix instead of a $4\times4$ one --- and for a
larger problem (say 47 regions with 7 lags) it is $4\times4$ blocks instead of
$47\times47$ ones, which is the difference between something you can read and
something you cannot.

```{r canada-latent}
lat_canada <- st_canada   # computed above
lat_canada
```

The convention is **row = effect at $t$, column = cause at $t-d$**: entry
$(q,q')$ says how strongly condition $q'$ in the previous quarter feeds
condition $q$ now. Diagonal entries are persistence, off-diagonal entries are
spillover between the latent conditions.

```{r canada-latent-parts}
# Persistence of each condition, and cross-condition spillover
diag(lat_canada$G$lag1)
lat_canada$G$lag1[1, 2]   # effect of Condition2 (t-1) on Condition1 (t)
```

The print-out also gives three stability numbers that say the same thing in
increasing order of sharpness:

* `colsum.max` --- $\max_j c_j$ of $\sum_d \Theta_d$. If it is below 1 the model
  is stationary, with no eigenvalue computation at all. It is only a sufficient
  condition, so a value above 1 (as here) is simply inconclusive.
* `spectral.radius.sum` --- $\rho(\sum_d G_d)$, which is below 1 exactly when
  the model is stationary.
* `spectral.radius` --- the spectral radius of the companion matrix itself.

The last one can be computed by either of two routes, selected with `method`.
The default `"latent"` uses the $DQ \times DQ$ companion matrix of the $G_d$;
`"companion"` uses the $PD \times PD$ companion matrix of the $\Xi_d$, as
earlier versions of the package always did. The two companion matrices share
their non-zero eigenvalues, so the answers agree --- but the latent route is a
$2\times2$ eigenproblem here instead of $4\times4$, and would be $28\times28$
instead of $329\times329$ for a 47-variable, 7-lag, rank-4 model.

```{r canada-latent-stability}
c(latent    = lat_canada$spectral.radius,
  companion = nmfkc.ar.stationarity(res_canada, method = "companion")$spectral.radius)
```

When the fit is stationary the process has a well-defined long-run mean, which
is reported in both the latent and the observed space ($\boldsymbol\mu_b$ and
$\boldsymbol\mu_y = X\boldsymbol\mu_b$):

```{r canada-latent-mean}
lat_canada$mu.b
lat_canada$mu.y
```

Two further quantities are useful when choosing the rank. `df` is the
*effective* dimension $Q(P+PD+1-Q)$ --- the right degrees of freedom when
comparing $R^2$ or information criteria across $Q$, since the naive count
$Q(P+PD+1)$ ignores the $Q^2$ rotational indeterminacy. `separability` is a
one-line identifiability check: values near 1 mean each basis owns an
essentially pure variable, so the factorization is unique up to permutation.

```{r canada-latent-df}
lat_canada$df
lat_canada$separability
```

### 4\. How Certain Is the Stability Verdict?

The $\rho$ reported above is below 1, so the fit is stationary --- but it is a
point estimate from 82 quarters, and the question that matters is whether it is
*reliably* below 1. `nmfkc.ar.stationarity.inference()` answers that by wild
bootstrap: it resamples the residuals, re-fits the whole NMF-VAR on each
replicate, and recomputes $\rho$.

```{r canada-inference}
inf_canada <- nmfkc.ar.stationarity.inference(st_canada, a_canada$Y, a_canada$A,
                                              wild.B = 200, wild.seed = 2026)
inf_canada
```

Two things are worth knowing about why this is bootstrapped rather than
delta-method'd.

**Only invariant quantities are bootstrapped.** The factorization is unique
only up to $(X,\Theta)\to(XT,T^{-1}\Theta)$, so an individual entry of $\Theta$
means nothing across replicates --- replicate 7's "Condition 1" need not be the
original's. Under that same map $G_d \to T^{-1}G_dT$, so the *spectrum* is
untouched: $\rho$, the eigenvalues, $\Xi_d$ and $\boldsymbol\mu_y$ are all
identified, and those are what the function reports. `p.star` (the composition
$X$ normalized by column) is invariant only up to basis *order*, so replicates
are permutation-aligned to the original before their quantiles are taken.

**The stationarity test is one-sided against $H_0:\rho\ge1$.** That is the
conservative direction: a small $p$ means the data rule out the unit root, and
the burden of proof is on stationarity rather than against it.

```{r canada-inference-parts}
c(rho = inf_canada$spectral.radius,
  se  = inf_canada$spectral.radius.se,
  lo  = inf_canada$spectral.radius.ci.lower,
  hi  = inf_canada$spectral.radius.ci.upper,
  p   = inf_canada$p.nonstationary)
```

The long-run mean gets an interval too, so the level the process reverts to can
be reported with uncertainty rather than as a bare number:

```{r canada-inference-mean}
rbind(estimate = inf_canada$mu.y,
      lower    = inf_canada$mu.y.ci.lower,
      upper    = inf_canada$mu.y.ci.upper)
```

Two diagnostics in the print-out deserve a look before the interval is trusted.
`truncation rate` is the fraction of replicate coefficients that hit the
non-negativity boundary and were clipped; a large value means the wild bootstrap
is being distorted by the constraint. `strict complementarity` checks the
boundary coordinates of the *original* fit: bootstrap consistency under an
active constraint requires the corresponding KKT multipliers to be strictly
positive, and a small `ratio` warns that a coefficient is sitting so close to
the boundary that replicates fall on both sides of it. The function warns when
either looks doubtful.

```{r canada-inference-diag}
c(truncate.rate = inf_canada$truncate.rate,
  n.boundary    = inf_canada$kkt$n.boundary,
  kkt.ratio     = inf_canada$kkt$ratio,
  usable.reps   = inf_canada$wild.B,
  failed.reps   = inf_canada$n.fail)
```

Each replicate is a full re-fit, so this is the expensive step of the analysis.
`wild.B = 200` is used here to keep the vignette quick; the default is 500, and
each re-fit converges to `epsilon = 1e-8` --- a looser tolerance leaves boundary
coefficients spuriously positive and understates the spread. On a multi-core
machine `cores = 4` divides the wall time without changing the answer.

### 5\. Latent Structure & Causal Graph

We can visualize how the two latent conditions change over time.

```{r canada-soft-cluster}
# Visualize soft clustering of time trends
barplot(res_canada$B.prob, col = c(2, 3), border = NA,
        main = "Soft Clustering of Economic Conditions",
        xlab = "Time", ylab = "Probability",
        names.arg = colnames(a_canada$Y), las=2, cex.names = 0.5)
legend("topright", legend = colnames(res_canada$X), fill = c(2, 3), bg = "white")
```

Finally, we can generate a **DOT script** to visualize the Granger causality (relationships) between variables inferred by the model.

```{r canada-dot}
# Generate DOT script for graph visualization
dot_script <- nmfkc.ar.DOT(res_canada, intercept = TRUE, threshold=0.01)

# plot(dot_script)  # requires DiagrammeR
```
