---
title: "Time Series Analysis with NMF-VAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Analysis with NMF-VAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{vars}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to apply the `nmfkc` package to time series data using the **NMF-VAR (Non-negative Matrix Factorization - Vector Autoregression)** framework. This approach models the coefficient matrix of NMF as a VAR process ($B \approx C A_{lag}$), allowing for both dimensionality reduction and temporal modeling.

Key functions in `nmfkc` such as `nmfkc.ar` and `nmfkc.ar.degree.cv` directly support R's native **`ts` objects**, preserving time series properties automatically. Once a model is fitted, `nmfkc.ar.stationarity()` checks dynamic stability, `nmfkc.ar.latent()` returns the $Q \times Q$ **latent transition matrices** $G_d = \Theta_d X$ --- which summarise the dynamics in the low-dimensional latent space rather than in the $P \times P$ observed space, and can be plotted directly --- and `nmfkc.ar.latent.inference()` attaches standard errors to them.

We will cover two scenarios:

1.  **Univariate AR:** Forecasting airline passengers (`AirPassengers`).
2.  **Multivariate VAR:** Analyzing macroeconomic indicators (`Canada`).

First, let's load the necessary packages.

```{r load-packages}
library(nmfkc)
library(vars) # For Canada dataset
```

-----

## Example 1: Univariate Autoregression with AirPassengers

The `AirPassengers` dataset contains monthly international airline passenger numbers. We will model this univariate time series using an autoregressive (AR) model.

### 1\. Data Preparation

We simply log-transform the `ts` object to stabilize variance. We do **not** need to manually convert it to a matrix or create time vectors; `nmfkc` functions handle `ts` objects directly.

```{r air-data-prep}
# Load and transform the ts object
d_air <- AirPassengers
d_air_log <- log10(d_air) # Still a ts object
```

### 2\. Model Selection (Lag Order)

We use **Cross-Validation** to find the optimal degree (lag order). By passing the `ts` object `d_air_log`, the function automatically handles the time dimension.

```{r air-degree-cv}
# Evaluate lag orders from 1 to 14
# Note: ts objects are automatically transposed to (Variables x Time) internally
cv_res <- nmfkc.ar.degree.cv(d_air_log, rank = 1, degree = 1:14, epsilon=1e-6, maxit=500000)

# Check the optimal degree
cv_res$degree

# For this example, we will proceed with D=12 (capturing monthly seasonality)
D <- 12
```

### 3\. Model Fitting

We construct the observation matrix `Y` and covariate matrix `A` (lagged Y) using `nmfkc.ar()`. The returned matrices `Y` and `A` will have time-based column names derived from the `ts` object.

```{r air-model-fit}
# Create matrices for the AR(12) model
a_air <- nmfkc.ar(d_air_log, degree = D, intercept = TRUE)

# Fit the NMF-AR model (Rank=1 for univariate)
res_air <- nmfkc(Y = a_air$Y, A = a_air$A, rank = 1, epsilon = 1e-6, maxit=500000)

# Check goodness of fit
res_air$r.squared

# Check for stationarity (spectral radius < 1)
st_air <- nmfkc.ar.stationarity(res_air)
c(rho = st_air$spectral.radius, stationary = st_air$stationary)
```

With `rank = 1` the latent VAR is one-dimensional, so the latent view is at its
simplest: the twelve $1\times1$ matrices $G_d$ are just the AR coefficients of
the single latent series, and `colsum.max` alone certifies stationarity.

```{r air-latent}
lat_air <- nmfkc.ar.latent(res_air)
c(rho = lat_air$spectral.radius, colsum.max = lat_air$colsum.max)
unlist(lat_air$G)[1:3]   # G_1, G_2, G_3 of the latent series
```

$\rho = 0.999$ is only just below 1, and the roots say why. All twelve sit close
to the unit circle at roughly equal angles: this is a seasonal root, not a slow
cycle. A single number cannot distinguish the two.

```{r air-roots, fig.height=4.4}
plot(lat_air, type = "roots")
```

The lag profile shows the same structure in the coefficients themselves ---
month 12 dominates, month 1 carries the short-run inertia.

```{r air-lag, fig.height=4.2}
plot(lat_air, type = "lag")
```

### 4\. Forecasting

We can forecast future values using the fitted model. `nmfkc.ar.predict` uses the time properties stored in the model to generate the correct future time sequence.

```{r air-forecast}
# Forecast next 2 years (24 months)
h <- 24
pred_res <- nmfkc.ar.predict(x = res_air, Y = a_air$Y, n.ahead = h)

# Convert predictions back to original scale
pred_val <- 10^as.vector(pred_res$pred)
pred_time <- pred_res$time # Future time points generated by the function

# --- Plotting ---
# Setup plot range
xlim_range <- range(c(time(d_air), pred_time))
ylim_range <- range(c(d_air, pred_val))

# 1. Observed data (Black)
plot(d_air, type = "l", col = "black", 
     xlim = xlim_range, ylim = ylim_range, lwd = 1,
     xlab = "Year", ylab = "Air Passengers", main = "NMF-VAR Forecast (h=24)")

# 2. Fitted values during training (Red)
# a_air$Y has column names as time strings; we parse them for plotting
fitted_time <- as.numeric(colnames(res_air$XB))
lines(fitted_time, 10^as.vector(res_air$XB), col = "red", lwd = 2)

# 3. Forecast (Blue)
# Connect the last observed point to the first forecast for a continuous line
last_t <- tail(as.numeric(time(d_air)), 1)
last_y <- tail(as.vector(d_air), 1)
lines(c(last_t, pred_time), c(last_y, pred_val), col = "blue", lwd = 2, lty = 2)

# Add legend
legend("topleft", legend = c("Observed", "Fitted", "Forecast"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = 2)
```

-----

## Example 2: Vector Autoregression with Canada Dataset

The `Canada` dataset contains four quarterly macroeconomic variables. We'll use a multivariate NMF-VAR model to analyze the relationship between these variables.

### 1\. Data Preparation

We take the first difference to achieve stationarity, normalize the data to $[0, 1]$, and transpose it to the `Variables x Time` format required by NMF.

```{r canada-data-prep}
# Load, difference, and normalize
d0_canada <- Canada
dd_canada <- apply(d0_canada, 2, diff) # Returns a matrix (Time x Vars)
dn_canada <- nmfkc.normalize(dd_canada)

# Transpose to (Variables x Time) for NMF
Y0_canada <- t(dn_canada)

# Create matrices for VAR(1)
a_canada <- nmfkc.ar(Y0_canada, degree = 1, intercept = TRUE)
```

### 2\. Model Fitting

We fit a model with `rank = 2` to identify two latent economic conditions driving the four variables.

```{r canada-model-fit}
# Fit the NMF-VAR model
# epsilon = 1e-8: this fit feeds the inference in Section 5, where coefficients
# sitting on the non-negativity boundary must actually have reached it.
res_canada <- nmfkc(Y = a_canada$Y, A = a_canada$A, rank = 2, prefix = "Condition", epsilon = 1e-8)

# R-squared and Stationarity
res_canada$r.squared
st_canada <- nmfkc.ar.stationarity(res_canada)
st_canada
```

The bracket says the cheap sufficient condition is inconclusive here, so the
eigenvalue is what decides.

```{r canada-bracket, fig.height=3.6}
plot(st_canada)
```

### 3\. Latent Transition Matrices

The fitted VAR acts on the four observed variables through the $P \times P$
matrices $\Xi_d = X\Theta_d$. Those are awkward to read, and they grow with the
number of series. But the model implies something much smaller: because
$\boldsymbol y_t = X\boldsymbol b_t$, the **coefficient** vector follows a
non-negative VAR of its own,

$$\boldsymbol b_t=\sum_{d=1}^{D}G_d\,\boldsymbol b_{t-d}+\boldsymbol\theta,
  \qquad G_d=\Theta_d X\ \ (Q\times Q).$$

`nmfkc.ar.latent()` returns these **latent transition matrices**. Here that
means a single $2\times2$ matrix instead of a $4\times4$ one --- and for a
larger problem (say 47 regions with 7 lags) it is $4\times4$ blocks instead of
$47\times47$ ones, which is the difference between something you can read and
something you cannot.

```{r canada-latent}
lat_canada <- nmfkc.ar.latent(res_canada)
lat_canada
```

The convention is **row = effect at $t$, column = cause at $t-d$**: entry
$(q,q')$ says how strongly condition $q'$ in the previous quarter feeds
condition $q$ now. Diagonal entries are persistence, off-diagonal entries are
spillover between the latent conditions.

```{r canada-latent-parts}
# Persistence of each condition, and cross-condition spillover
diag(lat_canada$G$lag1)
lat_canada$G$lag1[1, 2]   # effect of Condition2 (t-1) on Condition1 (t)
```

`nmfkc.ar.stationarity()` above reported three stability numbers that say the
same thing in increasing order of sharpness:

* `colsum.max` --- $\max_j c_j$ of $\sum_d \Theta_d$. If it is below 1 the model
  is stationary, with no eigenvalue computation at all. It is only a sufficient
  condition, so a value above 1 (as here) is simply inconclusive.
* `spectral.radius.sum` --- $\rho(\sum_d G_d)$, which is below 1 exactly when
  the model is stationary.
* `spectral.radius` --- the spectral radius of the companion matrix itself.

The last one can be computed by either of two routes, selected with `method`.
The default `"latent"` uses the $DQ \times DQ$ companion matrix of the $G_d$;
`"companion"` uses the $PD \times PD$ companion matrix of the $\Xi_d$, as
earlier versions of the package always did. The two companion matrices share
their non-zero eigenvalues, so the answers agree --- but the latent route is a
$2\times2$ eigenproblem here instead of $4\times4$, and would be $28\times28$
instead of $329\times329$ for a 47-variable, 7-lag, rank-4 model.

```{r canada-latent-stability}
c(latent    = lat_canada$spectral.radius,
  companion = nmfkc.ar.stationarity(res_canada, method = "companion")$spectral.radius)
```

When the fit is stationary the process has a well-defined long-run mean, which
is reported in both the latent and the observed space ($\boldsymbol\mu_b$ and
$\boldsymbol\mu_y = X\boldsymbol\mu_b$):

```{r canada-latent-mean}
lat_canada$mu.b
lat_canada$mu.y
```

Two further quantities are useful when choosing the rank. `df` is the
*effective* dimension $Q(P+PD+1-Q)$ --- the right degrees of freedom when
comparing $R^2$ or information criteria across $Q$, since the naive count
$Q(P+PD+1)$ ignores the $Q^2$ rotational indeterminacy. `separability` is a
one-line identifiability check: values near 1 mean each basis owns an
essentially pure variable, so the factorization is unique up to permutation.

```{r canada-latent-df}
lat_canada$df
lat_canada$separability
```

### 4\. Plotting the Latent Dynamics

Section 3 read the $G_d$ as numbers. `plot()` draws them. The point of every
panel below is that it stays readable at a size the observed space does not:
this fit has $Q^2 = 4$ latent impulse responses instead of $P^2 = 16$, and a
47-region, 7-lag, rank-4 fit would have 16 instead of 2209.

The **transition graph** shows the feedback structure. Self-loops are
persistence, arrows are spillover; a $\boldsymbol y\to\boldsymbol b\to
\boldsymbol y$ bipartite picture splits each round trip in half and so cannot
show that the loop is asymmetric.

```{r canada-plot-graph, fig.height=4.6}
plot(lat_canada, type = "graph")
```

The **roots on the unit circle** say *why* $\rho$ is where it is. Because the
$PD\times PD$ companion matrix has at most $DQ$ non-zero eigenvalues, all of
them fit on one plot. Complex roots are drawn separately and their period is
reported --- a fit can sit near the boundary because it cycles or because it is
seasonal, and a single number cannot tell those apart.

```{r canada-plot-roots, fig.height=4.6}
plot(lat_canada, type = "roots")
```

The **impulse responses** are non-negative for every horizon, since
$G_d\ge0$. The shape is therefore a decay (own effect) or a hump
(cross effect, zero at $h = 0$), and the half-life can be read straight off.

```{r canada-plot-irf, fig.height=5.4}
plot(lat_canada, type = "irf")
```

The **phase portrait** is the panel that changes an interpretation. Both
eigenvalues here are real and positive, so trajectories collapse along the fast
direction and then slide into $\boldsymbol\mu_b$ along the slow one: the fit is
a monotone relaxation, *not* a business cycle. Two coefficient series plotted
against time look "cyclical" either way; this plot does not.

```{r canada-plot-phase, fig.height=5.2}
plot(lat_canada, type = "phase")
```

One more panel is available: `type = "lag"` plots $(G_d)_{qq'}$ against $d$ ---
with $D = 1$ there is nothing to see, but for a multi-lag fit it answers "how
many periods does a shock take to arrive", inside the model rather than from
cross-correlations computed outside it. (The stationarity bracket is drawn by
`plot()` on the `nmfkc.ar.stationarity` object, as in Section 2.)

### 5\. Standard Errors for the Latent VAR

Every number in Sections 3 and 4 is a point estimate from 82 quarters.
`nmfkc.ar.latent.inference()` attaches a wild bootstrap to all of them: it
resamples the residuals, re-fits the whole NMF-VAR on each replicate, and
recomputes the latent quantities.

```{r canada-inference}
inf_canada <- nmfkc.ar.latent.inference(lat_canada, a_canada$Y, a_canada$A,
                                        wild.B = 200, wild.seed = 2026)
inf_canada
```

**What is identified decides what can be bootstrapped.** The factorization is
unique only up to $(X,\Theta)\to(XT,T^{-1}\Theta)$, so an individual entry of
$\Theta$ means nothing across replicates --- replicate 7's "Condition 1" need
not be the original's. Under that same map $G_d \to T^{-1}G_dT$, so the
*spectrum* is untouched: $\rho$, the eigenvalues and $\boldsymbol\mu_y$ are
invariant outright. The entries of $G_d$ and the composition
$\boldsymbol p^\ast$ are invariant only up to the *order* of the bases, and only
once the column scale of $X$ is pinned down --- which the default
`X.restriction = "colSums"` does. Each replicate is therefore permutation-aligned
to the original basis before its quantiles are taken, and the function warns if
the fit left the scale free.

The transition matrix now comes with standard errors, and the coefficient table
gives the same thing in long form:

```{r canada-inference-G}
inf_canada$coefficients
```

Read against the point estimates, the off-diagonal entries have $p$-values far
from 0 while the diagonal entries do not: the persistence of each condition is
solidly estimated, the spillover between them is not distinguishable from zero
at this sample size. The asymmetry seen in the graph is real in the point
estimate but not established by the data.

**The stationarity test is one-sided against $H_0:\rho\ge1$.** That is the
conservative direction: a small $p$ means the data rule out the unit root, and
the burden of proof is on stationarity rather than against it.

```{r canada-inference-parts}
c(rho = inf_canada$spectral.radius,
  se  = inf_canada$spectral.radius.se,
  lo  = inf_canada$spectral.radius.ci.lower,
  hi  = inf_canada$spectral.radius.ci.upper,
  p   = inf_canada$p.nonstationary)
```

The long-run mean gets an interval too, so the level the process reverts to can
be reported with uncertainty rather than as a bare number:

```{r canada-inference-mean}
rbind(estimate = inf_canada$mu.y,
      lower    = inf_canada$mu.y.ci.lower,
      upper    = inf_canada$mu.y.ci.upper)
```

Two diagnostics in the print-out deserve a look before the interval is trusted.
`truncation rate` is the fraction of replicate coefficients that hit the
non-negativity boundary and were clipped; a large value means the wild bootstrap
is being distorted by the constraint. `strict complementarity` checks the
boundary coordinates of the *original* fit: bootstrap consistency under an
active constraint requires the corresponding KKT multipliers to be strictly
positive, and a small `ratio` warns that a coefficient is sitting so close to
the boundary that replicates fall on both sides of it. The function warns when
either looks doubtful.

```{r canada-inference-diag}
c(truncate.rate = inf_canada$truncate.rate,
  n.boundary    = inf_canada$kkt$n.boundary,
  kkt.ratio     = inf_canada$kkt$ratio,
  usable.reps   = inf_canada$wild.B,
  failed.reps   = inf_canada$n.fail)
```

Each replicate is a full re-fit, so this is the expensive step of the analysis.
`wild.B = 200` is used here to keep the vignette quick; the default is 500, and
each re-fit converges to `epsilon = 1e-8` --- a looser tolerance leaves boundary
coefficients spuriously positive and understates the spread. On a multi-core
machine `cores = 4` divides the wall time without changing the answer.

### 6\. Latent Structure & Causal Graph

We can visualize how the two latent conditions change over time.

```{r canada-soft-cluster}
# Visualize soft clustering of time trends
barplot(res_canada$B.prob, col = c(2, 3), border = NA,
        main = "Soft Clustering of Economic Conditions",
        xlab = "Time", ylab = "Probability",
        names.arg = colnames(a_canada$Y), las=2, cex.names = 0.5)
legend("topright", legend = colnames(res_canada$X), fill = c(2, 3), bg = "white")
```

Finally, we can generate a **DOT script** to visualize the Granger causality (relationships) between variables inferred by the model.

```{r canada-dot}
# Generate DOT script for graph visualization
dot_script <- nmfkc.ar.DOT(res_canada, intercept = TRUE, threshold=0.01)

# plot(dot_script)  # requires DiagrammeR
```
